---
title: "Counting.genes.all.data"
author: "Diana.kisakye"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: yes
    toc_depth: 3
    toc_float: yes
    highlight: tango
    number_sections: no
    fig_caption: yes
    theme: sandstone
    code_folding: hide
---

<style>
h1 {background: darkblue;color: white;padding-left: 7px;}
h2 {color: darkblue;}
.code-folding-btn {display: none;}
</style>
  
<script>
  function show_span(id) {
    var x = document.getElementById(id);
    if (x.style.display === 'none') {
      x.style.display = 'inline';
    } else {
      x.style.display = 'none';
    }
  }
function myFunction(id) {
  var x = document.getElementById(id);
  if (x.style.display === 'none') {
    x.style.display = 'block';
  } else {
    x.style.display = 'none';
  }
}
</script>

```{r cleanup, warning=FALSE, message=FALSE, error=FALSE}
# CHECK R VERSION
 stopifnot(R.version.string == "R version 3.5.3 (2019-03-11)")

# CLEANUP
# Clear all variables:
rm(list=ls(all=TRUE))

# Unload current packages:
if (!is.null(names(utils::sessionInfo()[["otherPkgs"]]))) pacman::p_unload("all") 
```

```{r setup}
# KNITR DEFAULTS
knitr::opts_chunk$set(eval = TRUE, fig.height = 10)
```

## Load packages  
```{r packages, results="hide", warning=FALSE, message=FALSE, error=FALSE}
library(magrittr)
library(dplyr)
library(rstudioapi)
```

```{r setup_script, eval=TRUE, results='hide'}
# GET CURRENT SCRIPT NAME
(this.script <- rstudioapi::getActiveDocumentContext() %>% .$path %>% basename)
getwd()
list.files()
stopifnot(this.script != "")
```

## Data files  
### Upregulated genes  
```{r}

data.files.up <- list.files(path="data.out", pattern="genes\\.up\\.uniq.*", full.names = TRUE)
data.files.up
basename(data.files.up)
```

### Downregulated genes  
```{r}
data.files.down <- list.files(path="data.out", pattern="genes\\.down\\.uniq.*", full.names = T)
basename(data.files.down)
```

### Load data: upregulated genes    
```{r}
genes.up <- lapply(data.files.up, function(f) {
  scan(file = f, what = "character")
}) %>% 
  magrittr::set_names(basename(data.files.up))
```

```{r}
# class(genes.up) # "list"
# length(genes.up) # 9
# sapply(genes.up, class)
# sapply(genes.up, length)
```


```{r}
data.frame(File=names(genes.up),
            Gene.count=sapply(genes.up, length)) %>% 
  knitr::kable(., row.names = F)
```

### Load data: downregulated genes    
```{r}
genes.down <- lapply(data.files.down, function(f) {
  scan(file = f, what = "character")
}) %>% 
  magrittr::set_names(basename(data.files.down))
```

```{r}
data.frame(File=names(genes.down),
            Gene.count=sapply(genes.down, length)) %>% 
  knitr::kable(., row.names = F)
```

## Number of regulated genes:
### Number of upregulated genes  
```{r}
up.all <- unlist(genes.up) %>% toupper
# class(up.all) # "character"
length(up.all) # 2728
```

```{r}
table.up   <- table(up.all) %>% as.data.frame(., stringsAsFactors = FALSE) %>%
  set_colnames(c("Gene", "Count")) %>%
  arrange(desc(Count))
dim(table.up) # 1955    2    2
DT::datatable(table.up)
```

### Number of downregulated genes  
```{r}
down.all <- unlist(genes.down) %>% toupper
length(down.all) # 3136
```

```{r}
table.down <- table(down.all) %>% as.data.frame(.,stringsAsFactors = FALSE) %>%
  set_colnames(c("Gene", "Count")) %>%
  arrange(desc(Count))
DT::datatable(table.down)
```

```{r}
length(unique(table.up$Gene)) # 1955
```

```{r}
length(unique(table.down$Gene)) # 2215
```


```{r}
t.up.count <- table(table.up$Count) %>%
 as.data.frame(stringsAsFactors  = FALSE) %>%
  set_colnames(c("Count", "Freq")) 
knitr::kable(t.up.count)
# insert this as a table with all counts for both up and down regulated genes.
# extract gene symbols in 3 or more datasets and perform gene set enrichment analysis; enrichr
```

```{r}
t.down.count <- table(table.down$Count) %>%
 as.data.frame(stringsAsFactors  = FALSE) %>%
  set_colnames(c("Count", "Freq")) 
knitr::kable(t.down.count)
```

### Creating list of up-regulated genes with a count >= 3
```{r}
genes.up.min.3 <- table.up %>%
  filter(Count >= 3) %>%
  magrittr::use_series("Gene") %>% 
  unique %>% 
  sort
# genes.up.min.3
length(genes.up.min.3) # 151
head(genes.up.min.3) #"ABCC3"  "ADORA3" "AIF1"   "ALDH2"  "ANG"    "APOC1" 
```

### Creating list of down-regulated genes with a count >= 3
```{r}
genes.down.min.3 <- table.down %>%
  filter(Count >=3) %>%
  # .$Gene %>% 
  magrittr::use_series("Gene") %>% 
  unique %>% 
  sort 
head(genes.down.min.3) #  "ACSL1"   "ACSL5"   "ADM"     "ADORA2A" "ADTRP"   "AK4"    
length(genes.down.min.3) # 161
```


## Save results  
```{r}
out.file.up <- paste0("data.out/", this.script, ".up.min3.txt")
cat(genes.up.min.3, file = out.file.up, sep = "\n" )
cat("File saved:", out.file.up)
```

```{r}
out.file.down <- paste0("data.out/", this.script, ".down.min3.txt")
cat(genes.down.min.3, file = out.file.down, sep = "\n" )
cat("File saved:", out.file.down)
```

