---
title: "Enrichr.output.200.up.go.terms"
author: '9401'
date: "`r Sys.Date()`"
output:
  html_notebook:
    df_print: paged
    toc: yes
    highlight: tango
    number_sections: no
    fig_caption: yes
    theme: sandstone
    code_folding: hide
    toc_depth: 3
    toc_float: yes
---
  
<style>
h1 {background: darkblue;color: white;padding-left: 7px;}
h2 {color: darkblue;}
.code-folding-btn {display: none;}
</style>
  
<script>
  function show_span(id) {
    var x = document.getElementById(id);
    if (x.style.display === 'none') {
      x.style.display = 'inline';
    } else {
      x.style.display = 'none';
    }
  }
function myFunction(id) {
  var x = document.getElementById(id);
  if (x.style.display === 'none') {
    x.style.display = 'block';
  } else {
    x.style.display = 'none';
  }
}
</script>

```{r cleanup, warning=FALSE, message=FALSE, error=FALSE}
# CHECK R VERSION
# stopifnot(R.version.string == "R version 3.5.3 (2019-03-11)")

# CLEANUP
# Clear all variables:
rm(list=ls(all=TRUE))

# Unload current packages:
if (!is.null(names(utils::sessionInfo()[["otherPkgs"]]))) pacman::p_unload("all") 
```

```{r setup}
# KNITR DEFAULTS
knitr::opts_chunk$set(eval = TRUE, fig.height = 10)
```

### Load Packages
```{r packages, results="hide", warning=FALSE, message=FALSE, error=FALSE}
if (!require(stringr)) install.packages("stringr")
if (!require(rstudioapi)) install.packages("rstudioapi")
if (!require(tidyr)) install.packages("tidyr")
if (!require(tibble)) install.packages("tibble")
if (!require(readxl)) install.packages("readxl")
if (!require(purrr)) install.packages("purrr")
if (!require(RColorBrewer)) install.packages("RColorBrewer")
if (!require(devtools)) install.packages("devtools")
if (!require(magrittr)) install.packages("magrittr")
if (!require(dplyr)) install.packages("dplyr")
if (!require(scales)) install.packages("scales")
if (!require(knitr)) install.packages("knitr")
if (!require(kableExtra)) install.packages("kableExtra")

library(magrittr)
library(dplyr)
library(scales)
library(knitr)
library(kableExtra)
library(RankProd)
library(stringr)
```

```{r setup_script, eval=TRUE, results='hide'}
# GET CURRENT SCRIPT NAME
(this.script <- rstudioapi::getActiveDocumentContext() %>% .$path %>% basename)
getwd()
list.files()
stopifnot(this.script != "")
```

### Load Enrichr files
```{r}
getwd()
go.files<- list.files(path = "data.in/enrichr/go.tables", full.names = T)
go.files
cat(basename(go.files), sep="<br>\n")

```

```{r}
go.data.list <- lapply(go.files, function(f) {
  a <- read.table(f, sep = "\t", header = T, stringsAsFactors = F)
  a
})  %>%
  set_names(stringr::str_remove_all(basename(go.files), "\\_table.up.txt"))

sapply(go.data.list, class)
```

```{r}
data.frame(List.item = names(go.data.list),
           class = sapply(go.data.list, class),
           Columns = sapply(go.data.list, ncol),
           Rows = comma(sapply(go.data.list, nrow))) %>% 
   knitr::kable(., row.names = F, align=c("l", "l", "c", "c"), format = "html", escape = F) %>% 
   kableExtra::kable_styling("striped", full_width = T)

```
### Get enriched terms with Adj.P.value <= 0.01
```{r}
go.list <- lapply(seq_along(go.data.list), function(i, list.names) {
  df <- go.data.list[[i]]
  x <- dplyr::select(df,Term, Overlap,Adjusted.P.value) %>%
       dplyr::filter(Adjusted.P.value <= 0.01)
}, list.names = names(go.data.list)) %>% 
  set_names(names(go.data.list))

    
```

### Summarise the data frame
```{r}
data.frame(
  List.item = names(go.list),
  Class = sapply(go.list, function(x) {paste(class(x), collapse=",")}),
  Columns = sapply(go.list, ncol),
  Rows = sapply(go.list, nrow)
) %>% 
  knitr::kable(., row.names = F, align=c("l", "l", "c", "c"), format = "html", escape = F) %>% 
  kableExtra::kable_styling("striped", full_width = T)
```

```{r}
go.list.subset<- lapply(seq_along(go.list), function(i, list.names) {
  df <- go.list[[i]]
  x <- head(df,2)
},
  list.names = names(go.list)) %>% 
  set_names(names(go.list))

# go.list.subset
# names(go.list.subset)
```
### Merge data frames
```{r}
class(go.list.subset) #list

merged.df <- go.list.subset %>% 
  purrr::reduce(dplyr::full_join)

merged.df

merged.df.final <- data.frame(
  # Library = names(go.list.subset),
  Term = merged.df$Term,
  Overlap = merged.df$Overlap,
  Adjusted.P.value = merged.df$Adjusted.P.value
)

merged.df.final

```

### Save file
```{r}
out.file <- paste("data.out/",this.script, ".txt")
out.file
write.table(merged.df.final, file = out.file, sep ="\t")
```
