---
title: "Enrichment analysis for down-regulated genes"
author: "D. Kisakye"
date: "`r Sys.Date()`"
output:
  html_notebook:
    code_folding: hide
    fig_caption: yes
    highlight: tango
    number_sections: no
    theme: sandstone
    toc: yes
    toc_depth: 3
    toc_float: yes
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '3'
---
  
<style>
h1 {background: darkblue;color: white;padding-left: 7px;}
h2 {color: darkblue;}
.code-folding-btn {display: none;}
</style>
  
<script>
  function show_span(id) {
    var x = document.getElementById(id);
    if (x.style.display === 'none') {
      x.style.display = 'inline';
    } else {
      x.style.display = 'none';
    }
  }
function myFunction(id) {
  var x = document.getElementById(id);
  if (x.style.display === 'none') {
    x.style.display = 'block';
  } else {
    x.style.display = 'none';
  }
}
</script>
  

```{r cleanup, warning=FALSE, message=FALSE, error=FALSE}
# CHECK R VERSION
# stopifnot(R.version.string == "R version 3.5.3 (2019-03-11)")

# CLEANUP
# Clear all variables:
rm(list=ls(all=TRUE))

# Unload current packages:
if (!is.null(names(utils::sessionInfo()[["otherPkgs"]]))) pacman::p_unload("all") 
```

```{r setup}
# KNITR DEFAULTS
knitr::opts_chunk$set(eval = TRUE, fig.height = 10)
```

```{r packages, results="hide", warning=FALSE, message=FALSE, error=FALSE}
# LOAD PACKAGES:
library(magrittr)
library(dplyr)
```

```{r setup_script, eval=TRUE, results='hide'}
# GET CURRENT SCRIPT NAME
(this.script <- rstudioapi::getActiveDocumentContext() %>% .$path %>% basename)
getwd()
list.files()
stopifnot(this.script != "")
```

## Downregulated genes  
### Find files with downregulated genes  
```{r}
getwd()
data.file <- (path = "./meta_analyses/meta_analysis_rankprod/data_out/rank_product_01g.3NA.top.200.down.200.txt")
data.file
cat(basename(data.file))
gene.list.down<- scan(file = data.file, what = "character")

 # gene.list.down %>%
   # lapply(scan(what = "character")) %>%
   # cat(basename(gene.list.down))
 # gene.list.down
# "rank_product_01g.3NA.top.200.down.200.txt"
```


## Enrichr libraries  
### Find library files  
```{r results="asis"}
enrichr.files <- list.files("data.in/enrichr", pattern=".*txt$", full.names = T)
cat(basename(enrichr.files), sep="\n")
```


### Load libraries  
```{r results="hide"}

enrichr.libs.list <- lapply(enrichr.files, function(f) {
  readLines(con=f)
}) %>% 
  set_names(stringr::str_remove_all(basename(enrichr.files), "enrichr_|\\.txt$"))
class(enrichr.libs.list) # "list"
length(enrichr.libs.list) # 4
sapply(enrichr.libs.list, class) %>% as.matrix 
enrichr.libs.list
enrichr.libs.list<- grep(pattern = "^TRRUST", invert = T)

# ARCHS4_TFs_Coexp                  "character"
# ChEA_2016                         "character"
# Submissions_TF_Gene_Coocurrence   "character"
# TRRUST_Transcription_Factors_2019 "character"
# ARCHS4_TFs_Coexp                  1724
# ChEA_2016                          645
# Submissions_TF_Gene_Coocurrence   1721
# TRRUST_Transcription_Factors_2019  571
```

Convert to vector:  
```{r results="hide"}
enrichr.libs.v <- unlist(enrichr.libs.list) %>% unname
class(enrichr.libs.v) # "character"
length(enrichr.libs.v) # 4661


```

### Define dataset search terms  
```{r results="asis"}
search.terms <- c("^HHEX human tf ARCHS4 coexpression\t",
                  "^NR1H3 23393188 ChIP-Seq",
                  "^LYL1\t")
cat(search.terms, sep="\n")
```

Reality check: Can all search terms be found just once in vector of enrichr libs?  
```{r}
tmp <- sapply(search.terms, function(st) {any(grepl(st, enrichr.libs.v, ignore.case=T))}) %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column() %>% 
  set_colnames(c("Search.Term", "Found")) %>% 
  mutate(Matches = sapply(search.terms, function(st) {sum(grepl(st, enrichr.libs.v, ignore.case=F))}))
knitr::kable(tmp)
stopifnot(all(tmp$Found == TRUE))
stopifnot(all(tmp$Matches == 1))
rm(tmp)
```

### Extract enrichr gene sets  
```{r}
# Use a function for better readability:
extract.gene.symbols <- function(string) {
  a <- strsplit(string, split="\t")
  b <- a[[1]]
  c <- b[-c(1,2)]
  return(c)
}
```

```{r}
paste(names(enrichr.libs.list), search.terms)
```


```{r}
# Define names for gene set list:
search.terms.mod <- stringr::str_remove_all(search.terms, "\\^|\t")
# names(enrichr.libs.list)
#search.terms.mod
list.names <- paste(names(enrichr.libs.list), search.terms.mod, sep=" / ") %>% 
  stringr::str_replace_all(" ", "_")
# cat(list.names, sep="\n")


# Loop through the search terms and extract gene symbol lists:
enrichr.gene.sets <- lapply(search.terms, function(st) {
  lib.match <- grep(st, enrichr.libs.v, value=T)
  stopifnot(length(lib.match) == 1)
  gene.symbols <- extract.gene.symbols(lib.match) %>% toupper
  return(gene.symbols)
}) %>% 
  set_names(list.names)

rm(search.terms.mod, list.names)

class(enrichr.gene.sets) # "list"
## length(enrichr.gene.sets) # 4
# names(enrichr.gene.sets)

str(enrichr.gene.sets)
# List of 4
#[1] "list"
#List of 4
 #$ ARCHS4_TFs_Coexp_/_ETV3L_human_tf_ARCHS4_coexpression: chr [1:299] "RSAD2" "MX2" "TRAV12-1" "PML" ...
 #$ ChEA_2016_/_24523406                                 : chr [1:1182] "BTBD19" "DICER1" "RFX2" "PDYN" ...
 #$ Submissions_TF_Gene_Coocurrence_/_ETV3L              : chr [1:299] "IL1R1" "CADM1" "EBF1" "RSAD2" ...
 #$ TRRUST_Transcription_Factors_2019_/_RELA_human       : chr [1:301] "BAD" "CCK" "PSMB9" "COL1A2" ...
" 
```


## Get overlap with each of the gene sets down regulated by MTb + calculate p.value:  
p.value calculated by cummulative hypergeometric test
```{r}
#-------------------------------------------------------------------------------
# Function: hypergeometric test
#-------------------------------------------------------------------------------
hypergeo <- function(white.drawn, white, black, drawn, do.log=FALSE) {
  # Info: http://digitheadslabnotebook.blogspot.com/2011/02/using-r-for-introductory-statistics_21.html
  # dhyper(q, m, n, k, log = FALSE)
  #        q = number of successes; "white balls drawn" (here: number of genes that overlap)
  #        m + n = N ; N = total number of genes
  #        m = "white balls in urn"; e.g. total number of TF-bound genes
  #        n = "black balls in urn"; e.g. total number of genes NOT bound by the TF
  #        k = "number of balls drawn from urn"; sample size
  if (white < 1) {return(NA)}
  
  p <- phyper(white.drawn-1, white, black, drawn, lower.tail = FALSE, log.p=do.log) #why subract the (-1)?
  return(p)
} # end: hypergeo
```


```{r}
geo2r.gse34151.file <- "data.in/GSE34151.txt"
stopifnot(file.exists(geo2r.gse34151.file))
all.genes <- read.table(file=geo2r.gse34151.file, sep="\t", header=T, stringsAsFactors = F) %>% 
  magrittr::use_series("Gene.symbol")
all.genes <- all.genes[!all.genes %in% c(NA, "")]
all.genes <- sort(unique(toupper(all.genes)))
all.genes.count <- length(all.genes)
rm(geo2r.gse34151.file)
cat("To caluclate p.value need to know the total number of 'known' genes:", all.genes.count) # 20762
```


```{r}
# geneset.list:   list of downregulated genes for each dataset from GEO
# gene.symbols.v: vector of gene symbols from the enrichr libraries

get.stats.table <- function(geneset.list, gene.symbols.v, all.genes) {
  df <- lapply(seq_along(geneset.list), function(i) {
  study <- names(geneset.list)[i]
  # Vector of overlapping genes:
  overlap <- intersect(genes.down[[i]], gene.symbols.v)
  # Number of overlapping genes:
  overlap.count <- length(overlap)
  # p.value (hypergeometric test):
  white.balls.count <- length(intersect(gene.symbols.v, all.genes))
  black.balls.count <- all.genes.count - white.balls.count
  drawn.balls.count <- length(geneset.list[[i]])
  p <- hypergeo(white.drawn = overlap.count,
                white = white.balls.count,
                black = black.balls.count,
                drawn = drawn.balls.count)
  data.frame(TB.Study = study,
             Regulated.genes = drawn.balls.count,
             Overlap = overlap.count,
             p.value = p)
}) %>% 
  do.call(rbind, .) %>% 
  mutate(TB.Study = stringr::str_remove(TB.Study, "^genes.*.uniq\\.")) %>% 
  mutate(TB.Study = stringr::str_remove(TB.Study, "\\.txt$")) %>% 
  mutate(p.value = formatC(p.value)) %>% 
  arrange(TB.Study)
return(df)
}

```


### Stats: r names(enrichr.gene.sets)[1]
```{r}
get.stats.table(geneset.list=genes.down,
                gene.symbols.v=enrichr.gene.sets[[1]],
                all.genes=all.genes) %>% 
  knitr::kable(., align=c("l", "c", "c", "r"))
```


### Stats: r names(enrichr.gene.sets)[2]
```{r}
get.stats.table(geneset.list=genes.down,
                gene.symbols.v=enrichr.gene.sets[[2]],
                all.genes=all.genes) %>% 
  knitr::kable(., align=c("l", "c", "c", "r"))
```

### Stats: r names(enrichr.gene.sets)[3]
```{r}
get.stats.table(geneset.list=genes.down,
                gene.symbols.v=enrichr.gene.sets[[3]],
                all.genes=all.genes) %>% 
  knitr::kable(., align=c("l", "c", "c", "r"))
```

### Stats: r names(enrichr.gene.sets)[4]
```{r}
get.stats.table(geneset.list=genes.down,
                gene.symbols.v=enrichr.gene.sets[[4]],
                all.genes=all.genes) %>% 
  knitr::kable(., align=c("l", "c", "c", "r"))
```





## Session info  
<button class="button" onclick="myFunction('DIV_5')">Show/hide session info</button>
<div id="DIV_5" class="div_default_hide">

```{r print_date_and_time}
Sys.time()
```

```{r print_session_info}
# sessionInfo()
devtools::session_info()
```
</div>
  
<script>
  var divsToHide = document.getElementsByClassName("div_default_hide");
for(var i = 0; i < divsToHide.length; i++)
{
  divsToHide[i].style.display = 'none';
}
</script>
