---
title: "GSE103092.analysis"
author: "Diana.kisakye"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: yes
    toc_depth: 3
    toc_float: yes
    highlight: tango
    number_sections: no
    fig_caption: yes
    theme: sandstone
    code_folding: hide
---

  
<style>
h1 {background: darkblue;color: white;padding-left: 7px;}
h2 {color: darkblue;}
.code-folding-btn {display: none;}
</style>
  
<script>
  function show_span(id) {
    var x = document.getElementById(id);
    if (x.style.display === 'none') {
      x.style.display = 'inline';
    } else {
      x.style.display = 'none';
    }
  }
function myFunction(id) {
  var x = document.getElementById(id);
  if (x.style.display === 'none') {
    x.style.display = 'block';
  } else {
    x.style.display = 'none';
  }
}
</script>

```{r cleanup, warning=FALSE, message=FALSE, error=FALSE}
# CHECK R VERSION
 stopifnot(R.version.string == "R version 3.5.3 (2019-03-11)")

# CLEANUP
# Clear all variables:
rm(list=ls(all=TRUE))

# Unload current packages:
if (!is.null(names(utils::sessionInfo()[["otherPkgs"]]))) pacman::p_unload("all") 
```

```{r setup}
# KNITR DEFAULTS
knitr::opts_chunk$set(eval = TRUE, fig.height = 10)
```

## Load packages
```{r packages, results="hide", warning=FALSE, message=FALSE, error=FALSE}
library(magrittr)
library(dplyr)
library(tidyr)
library(biomaRt)
library(AnnotationDbi) 
# BiocManager::install("org.Hs.eg.db")
library(org.Hs.eg.db)
```


```{r setup_script, eval=TRUE, results='hide'}
# GET CURRENT SCRIPT NAME
(this.script <- rstudioapi::getActiveDocumentContext() %>% .$path %>% basename)
getwd()
list.files()
stopifnot(this.script != "")
```


## Load data file
```{r}
data.file <- "data.in/GSE103092.txt"
stopifnot(file.exists(data.file))
df <- read.table(data.file, header = T, sep = "\t", stringsAsFactors = F)
dim(df) # 57355     9
head(df)
```

## Summary of the variables
```{r}
sapply(df, class) %>% as.data.frame
```

## Do all probes have a proper GB accession id?
```{r}
# Are there any instances of a single space instead of a GB accesion? 
any(df$GB_ACC == " ") # FALSE
# Are there any instances of one or more spaces instead of a GB accession?
any(grepl(" +", df$GB_ACC)) # FALSE
# Are there any instances of EMPTY STRINGS instead of a GB accession?
any(df$GB_ACC =="") # TRUE
sum(df$GB_ACC =="") #19202
df2 <- filter(df, GB_ACC !="") # remove probes that don't have a proper gene bank accession id
nrow(df) #57355
nrow(df2)  #38153

```
## Are any GB accession IDs duplicated ?
```{r}
any(duplicated(df2$GB_ACC)) # TRUE
sum(duplicated(df2$GB_ACC)) # 7474
gb.acc.unique <- unique(df2$GB_ACC) # only probes with a unique gene bank accession id
length(gb.acc.unique) # 30679
#(sum(30679,7474)) "38153"
```
## Load biomart data
```{r, results='hide'}
ensembl = useMart("ensembl", dataset = "hsapiens_gene_ensembl" )
Mapped.ids <- getBM(attributes = c("refseq_mrna","hgnc_symbol"),
        filters = "refseq_mrna",
        values = df2$GB_ACC,
        mart = ensembl)
dim(Mapped.ids) # 24584     2
sapply(Mapped.ids, class) 
# refseq_mrna hgnc_symbol 
# "character" "character" 
print(head(Mapped.ids), row.names = F)
```

## How many GB accession IDs could be mapped using BioMart?
```{r}
all(df$GB_ACC %in% Mapped.ids$refseq_mrna)
length(df$GB_ACC) # 57355
sum(df$GB_ACC %in% Mapped.ids$refseq_mrna) # 24685 gene accession ids in the original gene expression data could be mapped to gene symbols, many of which are duplicates
```

## How many unique GB accession IDs could be mapped?
```{r}
sum(gb.acc.unique %in% Mapped.ids$refseq_mrna) #19794
pc <- sum(gb.acc.unique %in% Mapped.ids$refseq_mrna)/ length(gb.acc.unique) *100 # 64.5% unique ids could be mapped using the biomart database. 
pc <- round(pc, digits = 1)
paste0(pc, "%") #" 64.5%"
```
## How many unique GB IDs could not be mapped?
```{r}
acc.not.mapped <- setdiff(df$GB_ACC, Mapped.ids$refseq_mrna) 
length(acc.not.mapped) # 10886
tail(acc.not.mapped)
head(acc.not.mapped)
# "AY766452"     "XR_109632"    "AK130765"     "NM_020914"    "NM_001077493" "AY358259"
# "NR_038853" "NR_038945" "AL039002"  "XR_132554" "NR_024075" "BG202344" 
```

## Map more GB Accession IDs using annotation hub
```{r}

Mapped.ids.2 <- select(org.Hs.eg.db, keys = acc.not.mapped, columns = c("REFSEQ","SYMBOL"), keytype = "REFSEQ") 
Mapped.ids.3 <- Mapped.ids.2 %>%
  filter( !is.na(SYMBOL)) 
 
#dim(Mapped.ids.3) # 2976    2
head(Mapped.ids.3)

```

## Combine the two mapping tables with rbind  
```{r}
# When combining two data frames with rbind, the column names must be the same:
#names(Mapped.ids) # "refseq_mrna" "hgnc_symbol"
# names(Mapped.ids.3) # "REFSEQ" "SYMBOL"
names(Mapped.ids) <- names(Mapped.ids.3)
mapped.ids.all <- rbind(Mapped.ids, Mapped.ids.3)
#dim(mapped.ids.all) # 27560     2
any(duplicated(mapped.ids.all)) # TRUE
sum((duplicated(mapped.ids.all))) # 4732

```

## Remove duplicate rows
```{r}
mapped.ids.uniq <- distinct(mapped.ids.all)
dim(mapped.ids.uniq) # 22828     2
```

```{r}
sum(Mapped.ids$SYMBOL %in% mapped.ids.uniq$SYMBOL) # 24584
sum(Mapped.ids.3$SYMBOL %in% mapped.ids.uniq$SYMBOL) # 2976

```


## What percentage of GB accession IDs could now be mapped to gene symbols?  
```{r}
# length(unique(df2$GB_ACC)) # 30679
 sum(mapped.ids.uniq$REFSEQ %in% df2$GB_ACC) # 22828
paste0(round((sum(mapped.ids.uniq$REFSEQ %in% df2$GB_ACC)) / (length(unique(df2$GB_ACC))) * 100, 1), "%") %>% 
  cat(.) #74.4%
```

## Merge mapped ids with expression data  
```{r}
# names(df2)
# "ID"        "adj.P.Val" "P.Value"   "t"         "B"         "logFC"     "GB_ACC"    "SEQUENCE"  "SPOT_ID"
names(mapped.ids.uniq)
# "REFSEQ" "SYMBOL"
df3 <- merge(x=df2, y=mapped.ids.uniq, by.x = "GB_ACC", by.y = "REFSEQ", all.x = T, all.y = F)
names(df3)
# "GB_ACC"    "ID"        "adj.P.Val" "P.Value"   "t"         "B"         "logFC"     "SEQUENCE"  "SPOT_ID"   "SYMBOL"
```

## How many cases are  missing gene symbols?
```{r}
sum(is.na(df3$SYMBOL)) # 9121
```

## Keep only those rows with identifiable gene symbols
```{r}
df4 <- subset(df3, !is.na(SYMBOL)) %>% 
  dplyr::select(SYMBOL, adj.P.Val, P.Value, logFC)
nrow(df4) # 29102
```


## Up-regulated probes
```{r}
df.probes.up <- df4 %>%
  filter(adj.P.Val <= 0.01 ) %>%
  filter(logFC >=1 )
nrow(df.probes.up) #8
```

## Up-regulated probes associated with gene symbols
```{r}
genes.up <- df4 %>%
  filter(adj.P.Val <= 0.01) %>%
  filter(logFC >=1 ) %>% 
  distinct(SYMBOL)
head(genes.up)
nrow(genes.up) # 8


any(duplicated(genes.up$SYMBOL))  # FALSE


genes.up.uniq <- sort(genes.up$SYMBOL)
length(genes.up.uniq) # 8 unique genes
```

## Save file
```{r}
out.file.up <- paste0("data.out/", this.script, ".genes.up.uniq.txt")
out.file.up
cat(genes.up.uniq, file = out.file.up, sep = "\n")
cat("File saved:", out.file.up)

```

## Down-regulated probes
```{r}
df.probes.down <- df4 %>%
  filter(adj.P.Val <= 0.01) %>% 
  filter(logFC <= -1)
nrow(df.probes.down) #56
```

## Down-regulated probes associated with gene symbols
```{r}
genes.down <- df4 %>%
  filter(adj.P.Val <= 0.01) %>%
  filter(logFC <= -1) %>%
  distinct(SYMBOL)
nrow(genes.down) # 48

any(duplicated(genes.down$SYMBOL )) # FALSE
genes.down.uniq <- sort(genes.down$SYMBOL)
length(genes.down.uniq) #48
head(genes.down)

```
## Save file
```{r}
out.file.down <- paste0("data.out/", this.script, ".genes.down.uniq.txt")
out.file.down
cat(genes.down.uniq, file = out.file.down, sep = "\n")
cat("File saved:", out.file.down)

```


## Regulated genes as a function of parameters
```{r}
# Create empty matrices
p.value.cutoffs <- c(0.001, 0.005, 0.01, 0.05)
fc.cutoffs.up <- c(0.5, 1, 1.5, 2)
parameter.table.up <- matrix(rep(NA, (length(p.value.cutoffs) * length(fc.cutoffs.up))),
                                 ncol = length(fc.cutoffs.up),
                                 dimnames = list(
                                as.character(p.value.cutoffs),
                                as.character(fc.cutoffs.up)),
                                )
parameter.table.up


fc.cutoffs.down <- fc.cutoffs.up*-1
fc.cutoffs.down
parameter.table.down <- parameter.table.up %>%
  set_colnames(fc.cutoffs.down)
parameter.table.down
```

## Up-regulated genes as a function of parameters
```{r}

for (i in seq_along(fc.cutoffs.up)) {
  for (j in seq_along(p.value.cutoffs)) {
    parameter.table.up[i,j] <- df4 %>%
      filter(adj.P.Val <= p.value.cutoffs[i]) %>%
      filter(logFC >= fc.cutoffs.up[j]) %>%
      distinct(SYMBOL) %>%
      nrow
 }

}
parameter.table.up
```

## Down-regulated genes as a function of parameters
```{r}
for (i in seq_along(fc.cutoffs.down)) {
  for (j in seq_along(p.value.cutoffs)) {
    parameter.table.down[i,j] <- df4 %>%
      filter(adj.P.Val <= p.value.cutoffs[i]) %>%
      filter(logFC <= fc.cutoffs.down[j]) %>%
      distinct(SYMBOL) %>%
      nrow 
  }
  
}
parameter.table.down
```


```{r}

```

