---
title: "Data analysis markdown document for BCG.validation"
author: "Diana.kisakye"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: yes
    toc_depth: 3
    toc_float: yes
    highlight: tango
    number_sections: no
    fig_caption: yes
    theme: sandstone
    code_folding: hide
---

  
<style>
h1 {background: darkblue;color: white;padding-left: 7px;}
h2 {color: darkblue;}
.code-folding-btn {display: none;}
</style>
  
<script>
  function show_span(id) {
    var x = document.getElementById(id);
    if (x.style.display === 'none') {
      x.style.display = 'inline';
    } else {
      x.style.display = 'none';
    }
  }
function myFunction(id) {
  var x = document.getElementById(id);
  if (x.style.display === 'none') {
    x.style.display = 'block';
  } else {
    x.style.display = 'none';
  }
}
</script>

```{r cleanup, warning=FALSE, message=FALSE, error=FALSE}
# CHECK R VERSION
 stopifnot(R.version.string == "R version 3.5.3 (2019-03-11)")

# CLEANUP
# Clear all variables:
rm(list=ls(all=TRUE))

# Unload current packages:
if (!is.null(names(utils::sessionInfo()[["otherPkgs"]]))) pacman::p_unload("all") 
```

```{r setup}
# KNITR DEFAULTS
knitr::opts_chunk$set(eval = TRUE, fig.height = 10)
```

## Load packages
```{r packages, results="hide", warning=FALSE, message=FALSE, error=FALSE}
library(magrittr)
library(dplyr)
library(rstudioapi)
```


```{r setup_script, eval=TRUE, results='hide'}
# GET CURRENT SCRIPT NAME
(this.script <- rstudioapi::getActiveDocumentContext() %>% .$path %>% basename)
getwd()
list.files()
stopifnot(this.script != "")
```



## Load data file
```{r}

data.file <- "data.in/GSE108363.BCG.validation.txt"
stopifnot(file.exists(data.file))
df <- read.table(file = data.file, header = T, sep = "\t", stringsAsFactors = F)
dim(df) # 47323     8
head(df)
```

## Summary of the variables
```{r}
sapply(df, class) %>% as.data.frame
```

## Do all probes have proper gene symbols?
```{r}
# Are there any instances of a single space instead of a gene symbol? 
any(df$Gene.symbol == " ") # FALSE
# Are there any instances of one or more spaces used instead of a gene symbol?
any(grepl(" +", df$Gene.symbol)) # FALSE

# Are there any instances of EMPTY STRINGS instead of a gene symbol?
any(df$Gene.symbol =="") # TRUE
sum(df$Gene.symbol =="") # 16057
```

## Remove the probes that do not have a proper gene symbols
```{r}
df2 <- filter(df, Gene.symbol !="") 
nrow(df) # 47323
nrow(df2) # 31266


# proportion of those with missing symbols
pc <- sum(df$Gene.symbol == "")/ length(df$Gene.symbol) * 100
pc <- round(pc, digits = 1)
paste0(pc, "%") # "33.9%"
```

## Are any Gene symbols  duplicated ?
```{r}
any(duplicated(df2$Gene.symbol)) # TRUE
sum(duplicated(df2$Gene.symbol)) #  10504
```


```{r}
df3 <- subset(df2) %>% 
  dplyr::select(Gene.symbol, adj.P.Val, P.Value, logFC)
nrow(df3) # 31266
```


## Up-regulated probes
```{r}
df.probes.up <- df3 %>%
  filter(adj.P.Val <= 0.01 ) %>%
  filter(logFC >=1 )
nrow(df.probes.up) # 456
```

## Up-regulated probes associated with unique gene symbols
```{r}
genes.up <- df3  %>%
  filter(adj.P.Val <= 0.01) %>%
  filter(logFC >=1 ) %>%
  distinct(Gene.symbol)
nrow(genes.up) # 379

any(duplicated(genes.up$Gene.symbol)) # FALSE

genes.up.uniq<- sort(genes.up$Gene.symbol)
head(genes.up.uniq)
```

## Save File
```{r}
out.file.up <- paste0("data.out/", this.script, ".genes.up.uniq.txt")
out.file.up
cat(genes.up.uniq, file = out.file.up, sep = "\n")
cat("File saved:", out.file.up)
```


## Down-regulated probes

```{r}
df.probes.down <- df3 %>%
  filter(adj.P.Val <= 0.01) %>% 
  filter(logFC <= -1)
nrow(df.probes.down) # 258
```

## Down-regulated probes associated with unique gene symbols
```{r}
genes.down <- df3 %>%
  filter(adj.P.Val <= 0.01) %>%
  filter(logFC <= -1) %>%
  distinct(Gene.symbol)
nrow(genes.down) # 225

any(duplicated(genes.down$Gene.symbol)) # FALSE

genes.down.uniq<- sort(genes.down$Gene.symbol)
head(genes.down.uniq)


```


## Save file
```{r}
out.file.down <- paste0("data.out/", this.script, ".genes.down.uniq.txt")
out.file.down
cat(genes.down.uniq, file = out.file.down, sep = "\n")
cat("File saved:", out.file.down)

```


## Regulated genes as a function of parameters
```{r}
# Create empty matrices
p.value.cutoffs <- c(0.05, 0.01, 0.005,0.001)
fc.cutoffs.up <- c(0.5, 1, 1.5, 2)
parameter.table.up<- matrix(rep(NA, (length(p.value.cutoffs) * length(fc.cutoffs.up))),
                                 ncol = length(fc.cutoffs.up),
                                 dimnames = list(
                                as.character(p.value.cutoffs),
                                as.character(fc.cutoffs.up)),
                                )
parameter.table.up


fc.cutoffs.down <- fc.cutoffs.up*-1
fc.cutoffs.down
parameter.table.down <- parameter.table.up %>%
  set_colnames(fc.cutoffs.down)
parameter.table.down
```

## Up-regulated genes as a function of parameters
```{r}

for (i in seq_along(fc.cutoffs.up)) {
  for (j in seq_along(p.value.cutoffs)) {
    parameter.table.up[i,j] <- df3 %>%
      filter(adj.P.Val <= p.value.cutoffs[i]) %>%
      filter(logFC >= fc.cutoffs.up[j]) %>%
      distinct(Gene.symbol) %>%
      nrow
  }
}
parameter.table.up
```

## Down-regulated genes as a function of parameters
```{r}
for (i in seq_along(fc.cutoffs.down)) {
  for (j in seq_along(p.value.cutoffs)) {
    parameter.table.down[i,j] <- df3 %>% 
      filter(adj.P.Val <= p.value.cutoffs[i]) %>%
      filter(logFC <= fc.cutoffs.down[j]) %>%
      distinct(Gene.symbol) %>%
      nrow
  }
  
}
parameter.table.down
```

