---
title: "Meta analysis - rank product method for GSE108363"
author: "A. Nohturfft"
date: "`r Sys.Date()`"
output:
  html_notebook:
    df_print: paged
    toc: yes
    highlight: tango
    number_sections: no
    fig_caption: yes
    theme: sandstone
    code_folding: hide
    toc_depth: 3
    toc_float: yes
---
  
<style>
h1 {background: darkblue;color: white;padding-left: 7px;}
h2 {color: darkblue;}
.code-folding-btn {display: none;}
</style>
  
<script>
  function show_span(id) {
    var x = document.getElementById(id);
    if (x.style.display === 'none') {
      x.style.display = 'inline';
    } else {
      x.style.display = 'none';
    }
  }
function myFunction(id) {
  var x = document.getElementById(id);
  if (x.style.display === 'none') {
    x.style.display = 'block';
  } else {
    x.style.display = 'none';
  }
}
</script>

## Objective  

Perform meta analysis of TB microarray data using the Rank Product approach.  
Overview:  

1. Load fold change data generated with Geo2r  
2. Aggregate by gene symbol; where the same gene is represented by several probes, calculate the _**median**_ fold change  
3. Merge fold change data from all datasets into a single matrix  
4. Rank genes by fold change (convert fold changes to ranks)  
5. Plot rank correlations with Trellis plot  
6. Calculate rank products and p values using RankProd package  

  
```{r include=FALSE}
# About <style> chunk above:
# * provides custom formatting for level-1 and level-2 headers
# * hides the 'code' buttons next to every code chunk in the html output; leaves only one 'code' button at the top of the html notebook.

# About the <script> chunk at the top and bottom of the file:
# * allows to insert <div>'s whose content visbility is toggled with a button.
# * I use this to show/hide te output of sessionInfo() at the end of the script (see below)
```


```{r include=FALSE}
# RMARKDOWN HELP
# Rmarkdown:
# https://bookdown.org/yihui/rmarkdown/

# Chunk options:
# https://yihui.name/knitr/options
```

```{r cleanup, warning=FALSE, message=FALSE, error=FALSE}
# CHECK R VERSION
# stopifnot(R.version.string == "R version 3.4.3 (2017-11-30)")

# CLEANUP
# Clear all variables:
rm(list=ls(all=TRUE))

# Unload current packages:
# if (!is.null(names(utils::sessionInfo()[["otherPkgs"]]))) pacman::p_unload("all")
```

```{r setup}
# KNITR DEFAULTS
knitr::opts_chunk$set(eval = TRUE, fig.height = 10)
```

```{r packages, results="hide", warning=FALSE, message=FALSE, error=FALSE}
# LOAD PACKAGES:
if (!require(magrittr)) install.packages("magrittr")
library(magrittr)
if (!require(dplyr)) install.packages("dplyr")
library(dplyr)
if (!require(scales)) install.packages("scales")
library(scales)
if (!require(knitr)) install.packages("knitr")
library(knitr)
if (!require(kableExtra)) install.packages("kableExtra")
library(kableExtra)
if (!require(RankProd)) install.packages("RankProd")
library(RankProd)
if (!require(GGally)) install.packages("GGally")
library(GGally)
```

```{r setup_script, eval=TRUE, results='hide'}
# GET CURRENT SCRIPT NAME
(this.script <- rstudioapi::getActiveDocumentContext() %>% .$path %>% basename)
getwd()
list.files()
stopifnot(this.script != "")
```

## GEO2R data  


### Geo2r data files  
```{r}
geo2r.data.files <- list.files(path="./data.in", recursive = T, 
                               pattern="^GSE108363.*txt$", full.names = T)
geo2r.data.files
cat(basename(geo2r.data.files), sep="<br>\n")

```


```{r results="asis"}
# muarray.data.files <- c(geo2r.data.files, tailleux.data.files)
# muarray.data.files <- grep("GSE29731", muarray.data.files, invert = T, value=T)
# cat(basename(muarray.data.files), sep="<br>\n")
```

### Load Geo2r data  
```{r results="hide"}
geo2r.data.list <- lapply(geo2r.data.files, function(f) {
  a <- read.table(f, sep="\t", header=T, quote='"', stringsAsFactors = F)
  names(a)[grep("symbol", names(a), ignore.case = T)] <- "Gene.symbol"
  # names(a) <- stringr::str_remove(names(a), "^X\\.") 
  names(a) <- stringr::str_remove(names(a), "\\.$")
  a
}) %>% 
  set_names(stringr::str_remove_all(basename(geo2r.data.files), "^GSE108363.")) 
  # set_names(stringr::str_remove_all(basename(geo2r.data.files), "\\.txt")) 

#names(geo2r.data.list) <- stringr::str_remove(names(geo2r.data.list), "^X")
#names(geo2r.data.list)[grep("DC", names(geo2r.data.list))] <- "A-BUGS-23_DC"
#names(geo2r.data.list)[grep("MP", names(geo2r.data.list))] <- "A-BUGS-23_MP"

# class(geo2r.data.list) # "list"
#length(geo2r.data.list) # 4
#sapply(geo2r.data.list, class)
#sapply(geo2r.data.list, nrow) %>% unname # 47323 47323 47323 47323
names(geo2r.data.list)


```

```{r}
data.frame(List.item = names(geo2r.data.list),
           class = sapply(geo2r.data.list, class),
           Columns = sapply(geo2r.data.list, ncol),
           Rows = comma(sapply(geo2r.data.list, nrow)),
           Has.Gene.Symbols = sapply(geo2r.data.list, function(x) {any(grepl("Gene.symbol", names(x)))})
           # Unique.genes = sapply(geo2r.data.list, function(x) {length(unique(x$))})
           ) %>% 
  mutate(Has.Gene.Symbols = cell_spec(Has.Gene.Symbols, "html", color = ifelse(Has.Gene.Symbols == "TRUE", "black", "red"))) %>% 
  knitr::kable(., row.names = F, align=c("l", "l", "c", "c", "c"), format = "html", escape = F) %>% 
  kable_styling("striped", full_width = T)
```

```{r}
head(geo2r.data.list[[2]])
```


## Aggregate logFC by gene symbol  
```{r}
fold.change.list <- lapply(seq_along(geo2r.data.list), function(i, list.names) {
  df <- geo2r.data.list[[i]]
  x <- dplyr::select(df, Gene.symbol, logFC) %>% 
    dplyr::mutate(Gene.symbol = ifelse(Gene.symbol=="", NA, Gene.symbol)) %>% 
    tidyr::drop_na(.) %>% 
    dplyr::mutate(logFC = as.numeric(logFC)) %>% 
    # dplyr::mutate(Gene.symbol = stringr::str_remove_all(Gene.symbol, '\\"')) %>%  
    tidyr::drop_na(.) %>% 
    dplyr::group_by(Gene.symbol) %>% 
    dplyr::summarise(logFC = median(logFC), .groups="drop") %>% 
    tidyr::drop_na(.)
  names(x)[names(x) == "logFC"] <- list.names[i]
  x
}, list.names = names(geo2r.data.list)) %>% 
  set_names(names(geo2r.data.list))
names(fold.change.list) <- stringr::str_remove(names(fold.change.list), "_Geo2r.*$")
```


```{r}
data.frame(
  List.item = names(fold.change.list),
  Class = sapply(fold.change.list, function(x) {paste(class(x), collapse=",")}),
  Columns = sapply(fold.change.list, ncol),
  Rows = comma(sapply(fold.change.list, nrow)),
  Duplicated.genes = sapply(fold.change.list, function(x) {any(duplicated(x$Gene.symbol))})
) %>% 
  knitr::kable(., row.names = F, align=c("l", "l", "c", "c", "c"), format = "html", escape = F) %>% 
  kable_styling("striped", full_width = T)
```


## Combine all data frames  
### Merge data frames  
```{r}
merged.df <- fold.change.list %>% purrr::reduce(dplyr::full_join, by="Gene.symbol")
names(merged.df) <- stringr:: str_remove(names(merged.df), "_Geo2r") %>% 
  stringr::str_remove("_symbols$") %>% 
  stringr::str_replace("discovery", "Disc") %>% 
  stringr::str_replace("validation", "Val") %>%
  stringr::str_remove(".txt")

 dim(merged.df) # 210762     5
head(merged.df)
```

```{r results="asis"}
cat("Number of rows in merged data frame:<b>", comma(nrow(merged.df)), "</b><br>\n")
cat("Number of gene symbols in merged data frame:<b>", comma(length(unique(merged.df$Gene.symbol))), "</b><br>\n")
```

### Remove rows / genes with missing data  
```{r}
merged.df.filt <- merged.df %>% 
  dplyr::filter(complete.cases(.))
```

```{r results="asis"}
cat("Number of rows in merged data frame <i>after filtering</i>:<b>", comma(nrow(merged.df.filt)), "</b><br>\n") # br = newline in html
cat("Number of gene symbols in merged data frame <i>after filtering</i>:<b>", comma(length(unique(merged.df.filt$Gene.symbol))), "</b><br>\n")
```

```{r}
head(merged.df.filt, 20)
```

### Convert to matrix  
```{r}
merged.mx <- as.matrix(merged.df.filt[,-1])
rownames(merged.mx) <- merged.df.filt$Gene.symbol
merged.mx[1:4,1:4]
```

## Calculate ranks  
Examples for `rank()` function:  
```{r}
x <- seq(-5, 5)[-6]
x
set.seed(42)

y <- sample(x, 10)
y
data.frame(Random.Number = y,
           Rank.down = rank(y, ties.method = "first"),
           Rank.up = rank((y * -1), ties.method = "first")
           ) %>% 
  knitr::kable(., row.names = F, align=c("c", "c", "c"), format = "html", escape = F) %>% 
  kable_styling("striped", full_width = T)
# y
#  5  -6  -2  -1  -8   1   9 -10  -3   2   7   6  -7   0  10   4  -4  -5  -9   3
# rank(y)
# 16   5   9  10   3  12  19   1   8  13  18  17   4  11  20  15   7   6   2  14
# rank(y * -1)
#  5  16  12  11  18   9   2  20  13   8   3   4  17  10   1   6  14  15  19   7
```

```{r}
test.mx <- matrix(sample(x, 15,replace = T), nrow = 5)
test.mx
```

```{r}
apply(test.mx,2, rank)
#      [,1] [,2] [,3]
# [1,]  2.5  4.5    4
# [2,]  4.0  4.5    2
# [3,]  5.0  1.0    4
# [4,]  1.0  2.0    4
# [5,]  2.5  3.0    1

apply(test.mx, 2, rank, ties.method="first")
     # [,1] [,2] [,3]
# [1,]    2    4    3
# [2,]    4    5    2
# [3,]    5    1    4
# [4,]    1    2    5
# [5,]    3    3    1
apply(test.mx, 2, rank, ties.method = "max")
#     [,1] [,2] [,3]
#[1,]    3    5    5
#[2,]    4    5    2
#[3,]    5    1    5
#[4,]    1    2    5
#[5,]    3    3    1

apply(test.mx,2,rank, ties.method = "min")

 #    [,1] [,2] [,3]
#[1,]    2    4    3
#[2,]    4    4    2
#[3,]    5    1    3
#[4,]    1    2    3
#[5,]    2    3    1
```


### Ranks: downregulated  
```{r}
ranks.down.mx <- apply(merged.mx, 2, rank, ties.method="average")
ranks.down.mx[1:10,1:4]
 
    # BCG.Disc BCG.Val Mtb.Disc Mtb.Val
# A1BG        17518   10280  15602.5 10192.0
# A1BG-AS1     8673    3352  16359.0 10592.5
# A1CF        15864    8637   1858.0  7527.5
# A2M         20607   20602  20624.0 20728.0
# A2ML1       15673    1735  13080.0  7342.0
# A3GALT2     16177   12208  16821.0  5621.0
# A4GALT        859    1945  16752.0 11690.5
# A4GNT       11945   18811    955.5  3383.5
# AAAS        19463   18464  17707.5 18666.5
# AACS         5529    9951   5485.5  2571.0
```





### Ranks: upregulated  
```{r}
ranks.up.mx <- apply((merged.mx * -1), 2, rank, ties.method="average")
ranks.up.mx[1:10,1:4]

        # BCG.Disc BCG.Val Mtb.Disc Mtb.Val
# A1BG         3245   10483   5160.5 10571.0
# A1BG-AS1    12090   17411   4404.0 10170.5
# A1CF         4899   12126  18905.0 13235.5
# A2M           156     161    139.0    35.0
# A2ML1        5090   19028   7683.0 13421.0
# A3GALT2      4586    8555   3942.0 15142.0
# A4GALT      19904   18818   4011.0  9072.5
# A4GNT        8818    1952  19807.5 17379.5
# AAAS         1300    2299   3055.5  2096.5
# AACS        15234   10812  15277.5 18192.0
```


## Plot ranks  
```{r}
x <- ranks.down.mx[,1:4] %>% 
  as.data.frame %>% 
  tibble::rownames_to_column(var= "Gene") %>% 
  tidyr::pivot_longer(-c("Gene"), names_to = "Dataset", values_to = "Rank")
head(x)

```

```{r}
panel.cor <- function(x, y, digits = 3, prefix = "", cex.cor, ...)
{
     usr <- par("usr"); on.exit(par(usr))
     par(usr = c(0, 1, 0, 1))
     r <- cor(x, y)
     txt <- format(c(r, 0.123456789), digits = digits)[1]
     txt <- paste0("r=", txt)
     # if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
     # text(0.5, 0.5, txt, cex = cex.cor * r)
     text(0.5, 0.5, txt, cex=2.5, col="blue", font = 3)
}
pairs(merged.mx, col = alpha("black", 0.3), pch=18, upper.panel = panel.cor,
      text.panel = function(x,y,lab,cex,font) {text(x,y,lab, cex=3, font=2)})
```


```{r}
pairs(ranks.down.mx, col = alpha("black", 0.3), pch=18, upper.panel = panel.cor,
      text.panel = function(x,y,lab,cex,font) {text(x,y,lab, cex=3, font=2)})
```

## Calculate (geometric) mean ranks + p values for down-regulated genes 
```{r}
# browseVignettes("RankProd")
# help(package="RankProd")
# help(RP.advance, package="RankProd")
# help(RankProducts, package="RankProd")
# class (cl) = experimental conditions, e.g. control (0) versus drug (1)
# origin = study/lab/experiment, e.g. different GEO data series
```

```{r}
cl.down <- rep(1, ncol(ranks.down.mx))
cl.down # 1 1 1 1 
origin.down <- seq_len(ncol(ranks.down.mx)) 
origin.down # 1 2 3 4 
down.rank.prod <- RankProd::RankProducts(data=ranks.down.mx,
                                       cl=cl.down,
                                       gene.names = row.names(ranks.down.mx))
# class(down.rank.prod.sub) # "list"
 # length(down.rank.prod.sub) # 
```

```{r}
sapply(down.rank.prod, class) %>% as.matrix
# RPs               "matrix" 
# RPrank            "matrix" 
# pfp               "matrix" 
# pval              "matrix" 
# AveFC             "matrix" 
# groups            "numeric"
# RandomPairs_ranks "matrix" 
# nrep              "integer"
dim(down.rank.prod$RPs) # 20762   2
head(down.rank.prod$RPs)

         # class1 < class2 class1 > class2
# A1BG           13006.895       6560.6679
# A1BG-AS1        8423.611       9850.5364
# A1CF            6610.666      11040.4999
# A2M            20639.936        104.9481
# A2ML1           7147.798       9995.6769
# A3GALT2        11689.663       6956.4630
```

```{r}
down.rp.df <- as.data.frame(down.rank.prod$RPs) %>% 
  tibble::rownames_to_column(.) %>% 
  dplyr::select(1:2) %>% 
  set_colnames(c("Gene", "RP"))
head(down.rp.df)
```

```{r}
down.pval.df <- as.data.frame(down.rank.prod$pval) %>% 
  tibble::rownames_to_column(.) %>% 
  dplyr::select(1:2) %>% 
  set_colnames(c("Gene", "pval"))
head(down.pval.df)
```

### Correct for multiple hypothesis testing  
```{r}
down.rp.pval.df <- merge(x=down.rp.df, y=down.pval.df, by="Gene", all=T) %>% 
  dplyr::mutate(pval.adj = p.adjust(pval, method = "fdr"))
down.rp.pval.df
range(down.rp.pval.df$RP) #   1.778279 20759.249911
range(down.rp.pval.df$pval) #  4.795679e-16 1.000000e+00
nrow(subset(down.rp.pval.df, pval <= 0.05)) # 1632
nrow(subset(down.rp.pval.df, pval <= 0.01)) # 914

nrow(subset(down.rp.pval.df, pval.adj <= 0.05)) # 512
nrow(subset(down.rp.pval.df, pval.adj <= 0.01)) # 305
```

```{r}
subset(down.rp.pval.df, pval.adj <= 0.01) %>% 
  dplyr::arrange(RP)
```


```{r}
down.genes.sel.05 <- subset(down.rp.pval.df, pval.adj <= 0.05) %>% 
  use_series("Gene") %>% sort
class(down.genes.sel.05) # "character"
length(down.genes.sel.05) # 150
```

```{r}
cat(down.genes.sel.05, sep=", ")
```
## Create a dataframe of core significantly down-regulated genes (represented in all datasets)
```{r}
down.df.sel.05 <- as.data.frame(ranks.down.mx) %>% 
  tibble::rownames_to_column("Gene") %>% 
  dplyr::filter(Gene %in% down.genes.sel.05)
head(down.df.sel.05)
#dim(down.df.sel.05) # 512   5
```

```{r}
down.mx.sel.05 <- as.matrix(down.df.sel.05[,-1]) %>% 
  set_rownames(down.df.sel.05$Gene)
down.mx.sel.05[1:4,1:4]
```

```{r}
pairs(down.mx.sel.05,  col = alpha("black", 0.4), pch=19, cex=2,
      upper.panel = panel.cor,
      text.panel = function(x,y,lab,cex,font) {text(x,y,lab, cex=3, font=2)})
```


```{r message=FALSE, warning=FALSE}
out.file.down <- paste0(this.script, ".down.05.txt")
cat(down.genes.sel.05, sep="\n", file=out.file.down)
cat("File saved:", out.file.down) # Question about this: for my dissertation, interest would be in the core down regulated genes, meaning that my list would include only SDEs in all datasets.(down.df.sel.05$gene)
```

## Calculate geometric mean + p values for up regulated genes.
```{r}
cl.up <- rep(1, ncol(ranks.up.mx))
cl.up # 1 1 1 1 
origin.up<- seq_len(ncol(ranks.up.mx)) 
origin.up # 1 2 3 4 
up.rank.prod <- RankProd::RankProducts(data=ranks.up.mx,
                                       cl=cl.up,
                                       gene.names = row.names(ranks.up.mx))

```
```{r}
sapply(up.rank.prod, class) %>% as.matrix
# RPs               "matrix" 
# RPrank            "matrix" 
# pfp               "matrix" 
# pval              "matrix" 
# AveFC             "matrix" 
# groups            "numeric"
# RandomPairs_ranks "matrix" 
# nrep              "integer"
dim(up.rank.prod$RPs) # 20762   2
head(up.rank.prod$RPs)
         # class1 < class2 class1 > class2
# A1BG           6560.6679       13006.895
# A1BG-AS1       9850.5364        8423.611
# A1CF          11040.4999        6610.666
# A2M             104.9481       20639.936
# A2ML1          9995.6769        7147.798
# A3GALT2        6956.4630       11689.663

```

```{r}
up.rp.df <- as.data.frame(up.rank.prod$RPs) %>% 
  tibble::rownames_to_column(.) %>% 
  dplyr::select(1:2) %>% 
  set_colnames(c("Gene", "RP"))
head(up.rp.df)
```

```{r}
up.pval.df <- as.data.frame(up.rank.prod$pval) %>% 
  tibble::rownames_to_column(.) %>% 
  dplyr::select(1:2) %>% 
  set_colnames(c("Gene", "pval"))
head(up.pval.df)
```

### Correct for multiple hypothesis testing  
```{r}
up.rp.pval.df <- merge(x=up.rp.df, y=up.pval.df, by="Gene", all=T) %>% 
  dplyr::mutate(pval.adj = p.adjust(pval, method = "fdr"))
up.rp.pval.df
range(up.rp.pval.df$RP) #    3.08007 20760.74994
range(up.rp.pval.df$pval) # 1.66588e-14 1.00000e+00
nrow(subset(up.rp.pval.df, pval <= 0.05)) #  1900
nrow(subset(up.rp.pval.df, pval <= 0.01)) # 1122

nrow(subset(up.rp.pval.df, pval.adj <= 0.05)) # 709
nrow(subset(up.rp.pval.df, pval.adj <= 0.01)) # 366
```

```{r}
subset(up.rp.pval.df, pval.adj <= 0.01) %>% 
  dplyr::arrange(RP)
```



```{r}
up.genes.sel.05 <- subset(up.rp.pval.df, pval.adj <= 0.05) %>% 
  use_series("Gene") %>% sort
class(up.genes.sel.05) # "character"
length(up.genes.sel.05) # 709
```

```{r}
cat(up.genes.sel.05, sep=", ")
```

## Create a dataframe of core significantly up-regulated genes(represented in all datasets)
```{r}
up.df.sel.05 <- as.data.frame(ranks.up.mx) %>% 
  tibble::rownames_to_column("Gene") %>% 
  dplyr::filter(Gene %in% up.genes.sel.05)
head(up.df.sel.05)
# dim(down.df.sel.05) # 512   5

```

```{r}
up.mx.sel.05 <- as.matrix(up.df.sel.05[,-1]) %>% 
  set_rownames(up.df.sel.05$Gene)
up.mx.sel.05[1:4,1:4]
```

```{r}
pairs(up.mx.sel.05,  col = alpha("black", 0.4), pch=19, cex=2,
      upper.panel = panel.cor,
      text.panel = function(x,y,lab,cex,font) {text(x,y,lab, cex=3, font=2)})
```

# Save file
```{r message=FALSE, warning=FALSE}
out.file.up <- paste0(this.script, ".up.05.txt")
cat(up.genes.sel.05, sep="\n", file=out.file.up)
cat("File saved:", out.file.up) # Question about this: for my dissertation, interest would be in the core down regulated genes, meaning that my list would include only SDEs in all datasets.(down.df.sel.05$gene)
```
```

## Session info  
<button class="button" onclick="myFunction('DIV_5')">Show/hide session info</button>
<div id="DIV_5" class="div_default_hide">

```{r print_date_and_time}
Sys.time()
```

```{r print_session_info}
# sessionInfo()
devtools::session_info()
```
</div>
  
<script>
  var divsToHide = document.getElementsByClassName("div_default_hide");
for(var i = 0; i < divsToHide.length; i++)
{
  divsToHide[i].style.display = 'none';
}
</script>
