---
title: "Meta analysis - rank product method"
author: "A. Nohturfft"
date: "`r Sys.Date()`"
output:
  html_notebook:
    df_print: paged
    toc: yes
    highlight: tango
    number_sections: no
    fig_caption: yes
    theme: sandstone
    code_folding: hide
    toc_depth: 3
    toc_float: yes
---
  
<style>
h1 {background: darkblue;color: white;padding-left: 7px;}
h2 {color: darkblue;}
.code-folding-btn {display: none;}
</style>
  
<script>
  function show_span(id) {
    var x = document.getElementById(id);
    if (x.style.display === 'none') {
      x.style.display = 'inline';
    } else {
      x.style.display = 'none';
    }
  }
function myFunction(id) {
  var x = document.getElementById(id);
  if (x.style.display === 'none') {
    x.style.display = 'block';
  } else {
    x.style.display = 'none';
  }
}
</script>

## Objective  

Perform meta analysis of TB microarray AND RNA-seq data using the Rank Sum approach.  
Overview:  

1. Microarray data: load log(fold change) data that Diana generated with Geo2r  
2. RNA-seq data: load published tables  
3. Aggregate microarray data by gene symbol; where the same gene is represented by several probes, calculate the _**median**_ log(fold change)  
4. Merge fold change data from all datasets into a single matrix (**outer join**)  
5. Filter out genes with missing values (NA): keep only genes with a maximum number of missing values  
6. In each dataset rank genes by fold change (convert fold changes to ranks; NAs stay NAs; maximum ranks will vary by NA count)  
7. Replace NAs with average ranks in the other datasets  
8. Now rank again so number of ranks in each dataset is the same  
9. Calculate **rank sum** statistics and p values using RankProd package  
10. Generate heatmaps of top 100 regulated genes  
  
```{r include=FALSE}
# About <style> chunk above:
# * provides custom formatting for level-1 and level-2 headers
# * hides the 'code' buttons next to every code chunk in the html output; leaves only one 'code' button at the top of the html notebook.

# About the <script> chunk at the top and bottom of the file:
# * allows to insert <div>'s whose content visbility is toggled with a button.
# * I use this to show/hide te output of sessionInfo() at the end of the script (see below)
```


```{r include=FALSE}
# RMARKDOWN HELP
# Rmarkdown:
# https://bookdown.org/yihui/rmarkdown/

# Chunk options:
# https://yihui.name/knitr/options
```

```{r cleanup, warning=FALSE, message=FALSE, error=FALSE}
# CHECK R VERSION
# stopifnot(R.version.string == "R version 3.4.3 (2017-11-30)")

# CLEANUP
# Clear all variables:
rm(list=ls(all=TRUE))

# Unload current packages:
# if (!is.null(names(utils::sessionInfo()[["otherPkgs"]]))) pacman::p_unload("all")
```

```{r setup}
# KNITR DEFAULTS
knitr::opts_chunk$set(eval = TRUE, fig.height = 10)
```

```{r packages, results="hide", warning=FALSE, message=FALSE, error=FALSE}
# LOAD PACKAGES:
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!requireNamespace("ComplexHeatmap", quietly = TRUE)) BiocManager::install("ComplexHeatmap")
if (!require(stringr)) install.packages("stringr")
if (!require(rstudioapi)) install.packages("rstudioapi")
if (!require(tidyr)) install.packages("tidyr")
if (!require(tibble)) install.packages("tibble")
if (!require(readxl)) install.packages("readxl")
if (!require(purrr)) install.packages("purrr")
if (!require(RColorBrewer)) install.packages("RColorBrewer")
if (!require(devtools)) install.packages("devtools")
if (!require(magrittr)) install.packages("magrittr")
if (!require(dplyr)) install.packages("dplyr")
if (!require(scales)) install.packages("scales")
if (!require(knitr)) install.packages("knitr")
if (!require(kableExtra)) install.packages("kableExtra")
if (!requireNamespace("RankProd", quietly = TRUE)) BiocManager::install("RankProd")

library(magrittr)
library(dplyr)
library(scales)
library(knitr)
library(kableExtra)
library(RankProd)
```

```{r setup_script, eval=TRUE, results='hide'}
# GET CURRENT SCRIPT NAME
(this.script <- rstudioapi::getActiveDocumentContext() %>% .$path %>% basename)
getwd()
list.files()
stopifnot(this.script != "")
```

## Data: microarray studies    

### Geo2r data files  
```{r results="hide"}
geo2r.data.files <- list.files(path="../source_data", recursive = T, 
                               pattern="^GSE.*Geo2r.*txt$", full.names = T)
cat(basename(geo2r.data.files), sep="<br>\n")
```

```{r results="hide"}
tailleux.data.files <- list.files(path="../source_data", recursive = T,
                                  pattern="*tailleux.*txt", full.names = T)
cat(basename(tailleux.data.files), sep="<br>\n")

```

```{r results="asis"}
muarray.data.files <- c(geo2r.data.files, tailleux.data.files)
muarray.data.files <- grep("GSE29731", muarray.data.files, invert = T, value=T)
cat(basename(muarray.data.files), sep="<br>\n")
```

### Load Geo2r data  
```{r results="hide"}
geo2r.data.list <- lapply(muarray.data.files, function(f) {
  a <- read.table(f, sep="\t", header=T, quote='"', stringsAsFactors = F)
  names(a)[grep("symbol", names(a), ignore.case = T)] <- "Gene.symbol"
  names(a) <- stringr::str_remove(names(a), "^X\\.")
  names(a) <- stringr::str_remove(names(a), "\\.$")
  a
}) %>% 
  set_names(stringr::str_remove_all(basename(muarray.data.files), "\\.txt"))

names(geo2r.data.list) <- stringr::str_remove(names(geo2r.data.list), "^X")
names(geo2r.data.list)[grep("DC", names(geo2r.data.list))] <- "A-BUGS-23_DC"
names(geo2r.data.list)[grep("MP", names(geo2r.data.list))] <- "A-BUGS-23_MP"

# class(geo2r.data.list) # "list"
# length(geo2r.data.list) # 4
# sapply(geo2r.data.list, class) # "data.frame"
# sapply(geo2r.data.list, nrow) %>% unname # 29102 47231 24501 24501
```

```{r}
data.frame(List.item = names(geo2r.data.list),
           class = sapply(geo2r.data.list, class),
           Columns = sapply(geo2r.data.list, ncol),
           Rows = comma(sapply(geo2r.data.list, nrow)),
           Has.Gene.Symbols = sapply(geo2r.data.list, function(x) {any(grepl("Gene.symbol", names(x)))})
           # Unique.genes = sapply(geo2r.data.list, function(x) {length(unique(x$))})
           ) %>% 
  mutate(Has.Gene.Symbols = cell_spec(Has.Gene.Symbols, "html", color = ifelse(Has.Gene.Symbols == "TRUE", "black", "red"))) %>% 
  knitr::kable(., row.names = F, align=c("l", "l", "c", "c", "c"), format = "html", escape = F) %>% 
  kableExtra::kable_styling("striped", full_width = T)
```

```{r results="hide"}
# head(geo2r.data.list[[2]])
```


### Aggregate logFC by gene symbol  
```{r}
geo.fold.change.list <- lapply(seq_along(geo2r.data.list), function(i, list.names) {
  df <- geo2r.data.list[[i]]
  x <- dplyr::select(df, Gene.symbol, logFC) %>% 
    dplyr::mutate(Gene.symbol = ifelse(Gene.symbol=="", NA, Gene.symbol)) %>% 
    tidyr::drop_na(.) %>% 
    dplyr::mutate(logFC = as.numeric(logFC)) %>% 
    dplyr::mutate(Gene.symbol = stringr::str_remove_all(Gene.symbol, '\\"')) %>% 
    tidyr::drop_na(.) %>% 
    dplyr::group_by(Gene.symbol) %>% 
    dplyr::summarise(logFC = median(logFC), .groups="drop") %>% 
    tidyr::drop_na(.) %>% 
    as.data.frame(., stringsAsFactors=FALSE)
  # names(x)[names(x) == "logFC"] <- list.names[i]
  x
}, list.names = names(geo2r.data.list)) %>% 
  set_names(names(geo2r.data.list))
names(geo.fold.change.list) <- stringr::str_remove(names(geo.fold.change.list), "_Geo2r.*$")
```


```{r}
data.frame(
  List.item = names(geo.fold.change.list),
  Class = sapply(geo.fold.change.list, function(x) {paste(class(x), collapse=",")}),
  Columns = sapply(geo.fold.change.list, ncol),
  Rows = comma(sapply(geo.fold.change.list, nrow)),
  Duplicated.genes = sapply(geo.fold.change.list, function(x) {any(duplicated(x$Gene.symbol))})
) %>% 
  knitr::kable(., row.names = F, align=c("l", "l", "c", "c", "c"), format = "html", escape = F) %>% 
  kableExtra::kable_styling("striped", full_width = T)
```

## Data: RNA-seq studies  
```{r}
# <a href="https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE64179">GSE64179</a>
# https://pubmed.ncbi.nlm.nih.gov/26392366/
add.geo.hyperlink <- function(id) {
  a <- "<a href='https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc="
  b <- "'>"
  c <- "</a>"
  res <- paste0(a, id, b, id, c)
  res
}

add.pubmed.hyperlink <- function(id) {
  a <- "<a href='https://pubmed.ncbi.nlm.nih.gov/"
  b <- "/'>"
  c <- "</a>"
  res <- paste0(a, id, b, id, c)
  res
}
```


```{r}
rs.dataset.table <- data.frame(
  c=c("GSE64179", "26392366", "6", "Dendritic cells"),
  b=c("GSE67427", "26586179", "8", "Monocyte derived macrophages"),
  a=c("GSE148731", "32341411", "6", "M1, M2 macrophages")
) %>%
  t %>% 
  as.data.frame(., stringsAsFactors=F) %>% 
  set_colnames(c("Accession", "Pubmed", "Replicates", "Cell.Type")) %>% 
  mutate(Accession=add.geo.hyperlink(Accession)) %>% 
  mutate(Pubmed = add.pubmed.hyperlink(Pubmed))
knitr::kable(rs.dataset.table, row.names = F, escape=F, format = "html") %>% 
  kableExtra::kable_styling("striped", full_width = T)
```


### Load RNA-seq data: GSE64179  
```{r message=FALSE}
# list.files("../source_data/rna_seq/GSE64179")
rna.seq.data.files <- list()
rna.seq.data.files$GSE64179 <- "../source_data/rna_seq/GSE64179/Pacis_2015_TableS6_DEG.xlsx"
stopifnot(file.exists(rna.seq.data.files$GSE64179))
rna.seq.data.list <- list()
rna.seq.data.list$GSE64179 <- readxl::read_xlsx(path=rna.seq.data.files$GSE64179, range = "B3:C18818") %>% 
  set_colnames(c("Gene.symbol", "logFC")) %>% 
  mutate(Gene.symbol = toupper(Gene.symbol)) %>% 
  group_by(Gene.symbol) %>% 
  summarise(logFC = median(logFC), .groups="drop") %>% 
  as.data.frame(., stringsAsFactors=FALSE)
```


```{r}
# WITHOUT GROUPING / AGGREGATING:
# dim(rna.seq.data$GSE64179) # 18815     2
# sapply(rna.seq.data$GSE64179, class) # "character"   "numeric"
# range(rna.seq.data$GSE64179$logFC) # -9.426613 11.612170
# sum(rna.seq.data$GSE64179$logFC == 0) # 0
# sum(duplicated(rna.seq.data$GSE64179$Gene.symbol)) # 37
# dups <- rna.seq.data$GSE64179$Gene.symbol[duplicated(rna.seq.data$GSE64179$Gene.symbol)]
# subset(rna.seq.data$GSE64179, Gene.symbol %in% dups) %>% 
#   arrange(Gene.symbol, logFC)
# head(rna.seq.data$GSE64179)
```

```{r}
# WITH AGGREGATING: 
# dim(rna.seq.data$GSE64179) # 18776     2
# sapply(rna.seq.data$GSE64179, class) # "character"   "numeric"
# range(rna.seq.data$GSE64179$logFC) # -9.426613 11.612170
# sum(rna.seq.data$GSE64179$logFC == 0) # 0
# sum(duplicated(rna.seq.data$GSE64179$Gene.symbol)) # 0
head(rna.seq.data.list$GSE64179) %>% 
  knitr::kable(., row.names = F, format="html") %>% 
  kableExtra::kable_styling("striped", full_width = T)
```




### Load RNA-seq data: GSE67472
```{r}
rna.seq.data.files$GSE67472 <-"../source_data/rna_seq/GSE67427/GSE67427_table_s2.txt"
stopifnot(file.exists(rna.seq.data.files$GSE67472))
#rna.seq.data.list <-list()
rna.seq.data.list$GSE67472<- read.table(file = rna.seq.data.files$GSE67472, header = T, sep = "\t", stringsAsFactors = FALSE)  %>% 
#str(rna.seq.data.list$GSE67472) #data.frame':	12728 obs. of  140 variables:
  # head(rna.seq.data.list$GSE67472)
  
  # names(rna.seq.data.list$GSE67472)
  # select(matches(c("name", "Rv.18")))
 select(matches(c("name","Rv.18.logFC"), ignore.case = TRUE)) %>%
   # head(rna.seq.data.list$GSE67472)
  set_colnames(c("Gene.symbol", "logFC")) %>%
  mutate(Gene.symbol = toupper(Gene.symbol)) %>% 
  group_by(Gene.symbol) %>% 
  summarise(logFC = median(logFC), .groups="drop") %>% 
  as.data.frame(., stringsAsFactors=FALSE)
 head(rna.seq.data.list$GSE67472)
```

```{r}
# WITHOUT GROUPING / AGGREGATING:
dim(rna.seq.data.list$GSE67472) #  12724     2
sapply(rna.seq.data.list$GSE67472, class) # "character"   "numeric"
range(rna.seq.data.list$GSE67472$logFC) # -4.155479  8.205439
sum(rna.seq.data.list$GSE67472$logFC == 0) # 0
sum(duplicated(rna.seq.data.list$GSE67472$Gene.symbol)) # 0
# dups <- rna.seq.data$GSE64179$Gene.symbol[duplicated(rna.seq.data$GSE64179$Gene.symbol)]
# subset(rna.seq.data$GSE64179, Gene.symbol %in% dups) %>% 
#   arrange(Gene.symbol, logFC)
# head(rna.seq.data$GSE64179)
```

```{r}
# WITH AGGREGATING: 
# dim(rna.seq.data$GSE64179) # 18776     2
# sapply(rna.seq.data$GSE64179, class) # "character"   "numeric"
# range(rna.seq.data$GSE64179$logFC) # -9.426613 11.612170
# sum(rna.seq.data$GSE64179$logFC == 0) # 0
# sum(duplicated(rna.seq.data$GSE64179$Gene.symbol)) # 0
head(rna.seq.data.list$GSE67472) %>% 
  knitr::kable(., row.names = F, format="html") %>% 
  kableExtra::kable_styling("striped", full_width = T)

```

### Load RNA-seq data: GSE48731.MF1
```{r message=FALSE}
# list.files("../source_data/rna_seq/GSE148731")
# rna.seq.data.files <- list()
rna.seq.data.files$GSE148731.MF1 <- "../source_data/rna_seq/GSE148731/GSE148731_toptable_MF1_01.Rmd.2020_07_09.txt"
stopifnot(file.exists(rna.seq.data.files$GSE148731.MF1))
 # rna.seq.data.list <- list()
rna.seq.data.list$GSE148731.MF1 <- read.table(file = rna.seq.data.files$GSE148731.MF1, header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>%
   select(c("Gene.symbol", "logFC")) %>% 
  # mutate(Gene.symbol = toupper(Gene.symbol)) %>% 
  # summarise(logFC = median(logFC), .groups="drop") %>% 
   as.data.frame(., stringsAsFactors=FALSE)
head(rna.seq.data.list$GSE148731.MF1)
```
```{r}
# WITH AGGREGATING:
dim(rna.seq.data.list$GSE148731.MF1) # 17802     2
sapply(rna.seq.data.list$GSE148731.MF1, class) # "character"   "numeric"
range(rna.seq.data.list$GSE148731.MF1$logFC) # -2.386124  9.704865
sum(rna.seq.data.list$GSE148731.MF1$logFC == 0) #
sum(duplicated(rna.seq.data.list$GSE148731.MF1$Gene.symbol)) # 0

```

```{r}
head(rna.seq.data.list$GSE148731.MF1) %>% 
  knitr::kable(., row.names = F, format="html") %>% 
  kableExtra::kable_styling("striped", full_width = T)
```


### Load RNA-seq data: GSE48731.MF2
```{r message=FALSE}
 # list.files("../source_data/rna_seq/GSE148731")
# rna.seq.data.files <- list()
rna.seq.data.files$GSE148731.MF2 <- "../source_data/rna_seq/GSE148731/GSE148731_toptable_MF2_01.Rmd.2020_07_09.txt"
stopifnot(file.exists(rna.seq.data.files$GSE148731.MF2))
 # rna.seq.data.list <- list()
rna.seq.data.list$GSE148731.MF2 <- read.table(file = rna.seq.data.files$GSE148731.MF2, header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>%
   select(c("Gene.symbol", "logFC")) %>% 
  # mutate(Gene.symbol = toupper(Gene.symbol)) %>% 
  # summarise(logFC = median(logFC), .groups="drop") %>% 
   as.data.frame(., stringsAsFactors=FALSE)
head(rna.seq.data.list$GSE148731.MF2)
```


```{r}
# WITH AGGREGATING:
dim(rna.seq.data.list$GSE148731.MF2) #  17085     2
sapply(rna.seq.data.list$GSE148731.MF2, class) # "character"   "numeric"
range(rna.seq.data.list$GSE148731.MF2$logFC) # -7.524136  9.795100
sum(rna.seq.data.list$GSE148731.MF2$logFC == 0) #
sum(duplicated(rna.seq.data.list$GSE148731.MF2$Gene.symbol)) # 0

```

```{r}
head(rna.seq.data.list$GSE148731.MF2) %>% 
  knitr::kable(., row.names = F, format="html") %>% 
  kableExtra::kable_styling("striped", full_width = T)
```

## Combine all data frames  
```{r}
# length(rna.seq.data.list) #4
# length(geo.fold.change.list)# 4
all.data.list <- c(geo.fold.change.list, rna.seq.data.list)
all(sapply(all.data.list, class) == "data.frame") # TRUE
```

```{r}
lapply(all.data.list, names)
length(all.data.list) # 8
```


### Merge data frames  
```{r}
merged.df <- all.data.list %>% purrr::reduce(dplyr::full_join, by="Gene.symbol") %>% 
  tibble::column_to_rownames("Gene.symbol") %>% 
  set_colnames(paste0("dataset", seq_along(.)))

dim(merged.df) # 30687     8
head(merged.df) %>% 
  knitr::kable(., row.names = T, format="html") %>% 
  kableExtra::kable_styling("striped", full_width = T)
```

Summary table:  
```{r}
df.summary <- data.frame(
  Column = names(merged.df),
  Dataset=names(all.data.list),
  Min = round(sapply(merged.df, min, na.rm=T), 2),
  Max = round(sapply(merged.df, max, na.rm=T), 2),
  Count.zero = sapply(merged.df, function(x) {sum(x==0, na.rm=T)}),
  Count.NA = comma(sapply(merged.df, function(x) {sum(is.na(x))}))
) %>% 
  mutate(Count.zero = comma(as.integer(Count.zero), accuracy=1))
knitr::kable(df.summary, row.names = F, format="html", align=rep("l", 5)) %>% 
  kableExtra::kable_styling("striped", full_width = T)
```


```{r results="asis"}
cat("Number of rows in merged data frame:<b>", comma(nrow(merged.df)), "</b><br>\n")
cat("Number of gene symbols in merged data frame:<b>", comma(length(unique(row.names(merged.df)))), "</b><br>\n")
```

```{r results="asis"}
cat("Number of complete data rows: <b>", comma(sum(complete.cases(merged.df))), "</b> (",
    percent(sum(complete.cases(merged.df))/length(unique(row.names(merged.df)))),
    ")<br>\n", sep="")
```

### Filter rows: One NA max  
```{r}
one.na.max.count <- as.matrix(merged.df) %>% 
  apply(., 1, function(x) {  # 1 = select rows
    n <- sum(is.na(x))
    n
  })
 class(one.na.max.count) # integer
length(one.na.max.count) # 30687
merged.df.filt.na <- merged.df[one.na.max.count <= 1,]
dim(merged.df.filt.na)   # 8626    8
 dim(merged.df)    #  30687     8  
```

Print 10 rows:  
```{r}
head(merged.df.filt.na, 10) %>% 
  knitr::kable(., row.names = T, format="html", align=rep("l", 5)) %>% 
  kableExtra::kable_styling("striped", full_width = T)
```

### Summary table after filtering NAs:  
```{r}
df.summary.2 <- data.frame(
  Column = names(merged.df.filt.na),
  Dataset=names(all.data.list),
  Min = round(sapply(merged.df.filt.na, min, na.rm=T), 2),
  Max = round(sapply(merged.df.filt.na, max, na.rm=T), 2),
  Count.zero = sapply(merged.df.filt.na, function(x) {sum(x==0, na.rm=T)}),
  Count.NA = comma(sapply(merged.df.filt.na, function(x) {sum(is.na(x))}))
) %>% 
  mutate(Count.zero = comma(as.integer(Count.zero), accuracy=1))
knitr::kable(df.summary.2, row.names = F, format="html", align=rep("l", 5)) %>% 
  kableExtra::kable_styling("striped", full_width = T)
rm(df.summary.2)
```



## Calculate ranks (1)  

Key paramters:  

* NAs stay NAs  
* ties are averaged  
* keeping NAs results in different maximum ranks  
* (will have to rank again after filling NAs)  

### Ranks: downregulated  
```{r}
ranks.down.mx <- apply(merged.df.filt.na, 2, rank, ties.method="average", na.last="keep")
# class(ranks.down.mx) # matrix
# dim(ranks.down.mx) # 12786     5
# ranks.down.mx[1:10,]
```

Print 6 rows (sorted):  
```{r}
ranks.down.mx[order(rowMeans(ranks.down.mx, na.rm=T)),] %>% head
```

The maximum ranks will vary by dataset:  
```{r}
apply(ranks.down.mx, 2, max, na.rm=T) %>%
  as.data.frame(., stringsAsFactors=F) %>% 
  tibble::rownames_to_column() %>% 
  set_colnames(c("Dataset (down)", "Max.Rank")) %>% 
  mutate(Max.Rank = comma(Max.Rank)) %>% 
  knitr::kable(., row.names = F, align=c("l","r"), format="html") %>% 
  kableExtra::kable_styling("striped")
```

### Ranks: upregulated  
```{r}
ranks.up.mx <- apply((merged.df.filt.na * -1), 2, rank, ties.method="average", na.last="keep")
```

Print 6 rows (sorted):  
```{r}
ranks.up.mx[order(rowMeans(ranks.up.mx, na.rm=T)),] %>% head
```


```{r}
apply(ranks.up.mx, 2, max, na.rm=T) %>%
  as.data.frame(., stringsAsFactors=F) %>% 
  tibble::rownames_to_column() %>% 
  set_colnames(c("Dataset (up)", "Max.Rank")) %>% 
  mutate(Max.Rank = comma(Max.Rank)) %>% 
  knitr::kable(., row.names = F, align=c("l","r"), format="html") %>% 
  kableExtra::kable_styling("striped")
```

## Replace missing values with averages  
```{r}
na.to.average <- function(x) {
  any.na <- any(is.na(x))
  if (isTRUE(any.na)) {
    indx.na <- which(is.na(x))
    ave <- mean(x, na.rm=T)
    x[indx.na] <- ave
  }
  x
}
```

### Replace NAs: downregulation  
```{r}
ranks.down.nona.mx <- t(apply(ranks.down.mx, 1, na.to.average))
# dim(ranks.down.mx) # 12786     5
# dim(ranks.down.nona.mx) # 12786     5
# sum(is.na(ranks.down.mx)) # 1436
# sum(is.na(ranks.down.nona.mx)) # 0
# ranks.down.nona.mx[1:4,1:4]
ranks.down.nona.mx[order(rowMeans(ranks.down.nona.mx, na.rm=T)),] %>% 
  head
```

### Replace NAs: upregulation  
```{r}
ranks.up.nona.mx <- t(apply(ranks.up.mx, 1, na.to.average))
# dim(ranks.up.mx) # 12786     5
# dim(ranks.up.nona.mx) # 12786     5
# sum(is.na(ranks.up.mx)) # 1436
# sum(is.na(ranks.up.nona.mx)) # 0
# ranks.up.nona.mx[1:4,1:4]
ranks.up.nona.mx[order(rowMeans(ranks.up.nona.mx, na.rm=T)),] %>% 
  head
```


## Calculate ranks (2)  
Having replaced missing values with averages, repeat the ranking such that maximum ranks will be the same for each dataset.  
Key paramters:  

* ties are averaged  

### Ranks: downregulated (2)  
```{r}
ranks.down.mx.2 <- apply(ranks.down.nona.mx, 2, rank, ties.method="average", na.last="keep")
# class(ranks.down.mx.2) # matrix
# dim(ranks.down.mx.2) # 12786     5
# ranks.down.mx.2[1:10,]
ranks.down.mx.2[order(rowMeans(ranks.down.mx.2, na.rm=T)),] %>% 
  head
```

```{r}
apply(ranks.down.mx.2, 2, max) %>%
  as.data.frame(., stringsAsFactors=F) %>% 
  tibble::rownames_to_column() %>% 
  set_colnames(c("Dataset (down)", "Max.Rank")) %>% 
  knitr::kable(., row.names = F, align=c("l","r"), format="html") %>% 
  kableExtra::kable_styling("striped")
```

### Ranks: upregulated (2)  
```{r}
ranks.up.mx.2 <- apply(ranks.up.nona.mx, 2, rank, ties.method="average", na.last="keep")
# ranks.up.mx.2[1:10,]
ranks.up.mx.2[order(rowMeans(ranks.up.mx.2, na.rm=T)),] %>% head
```

```{r}
apply(ranks.up.mx.2, 2, max) %>%
  as.data.frame(., stringsAsFactors=F) %>% 
  tibble::rownames_to_column() %>% 
  set_colnames(c("Dataset (up)", "Max.Rank")) %>% 
  knitr::kable(., row.names = F, align=c("l","r"), format="html") %>% 
  kableExtra::kable_styling("striped")
```



## Plot correlations  
### Plot fold changes  
```{r}
panel.cor <- function(x, y, digits = 3, prefix = "", cex.cor, ...)
{
     usr <- par("usr"); on.exit(par(usr))
     par(usr = c(0, 1, 0, 1))
     r <- cor(x, y, use="pairwise.complete.obs")
     txt <- format(c(r, 0.123456789), digits = digits)[1]
     txt <- paste0("r=", txt)
     # if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
     # text(0.5, 0.5, txt, cex = cex.cor * r)
     text(0.5, 0.5, txt, cex=2.5, col="blue", font = 3)
}
merged.df %>% 
  set_colnames(names(all.data.list)) %>% 
  pairs(., col = alpha("black", 0.3), pch=18, upper.panel = panel.cor,
        xaxt='n', yaxt='n',
        text.panel = function(x,y,lab,cex,font) {text(x,y,lab, cex=3, font=2)})
```

### Plot ranks  
```{r}
panel.cor.spearman <- function(x, y, digits = 3, prefix = "", cex.cor, ...)
{
     usr <- par("usr"); on.exit(par(usr))
     par(usr = c(0, 1, 0, 1))
     r <- cor(x, y, use="pairwise.complete.obs", method="spearman")
     txt <- format(c(r, 0.123456789), digits = digits)[1]
     txt <- paste0("r=", txt)
     text(0.5, 0.5, txt, cex=2.5, col="purple", font = 3)
}
ranks.down.mx.2 %>% 
  set_colnames(names(all.data.list)) %>% 
  pairs(., col = alpha("black", 0.3), pch=18, upper.panel = panel.cor.spearman,
        xaxt='n', yaxt='n',
      text.panel = function(x,y,lab,cex,font) {text(x,y,lab, cex=3, font=2)})
```

## Calculate rank sum statistics  
```{r}
# browseVignettes("RankProd")
# help(package="RankProd")
# help(RankProducts, package="RankProd")
```

```{r}
get.rank.sum <- function(mx) {
  cl.down <- rep(1, ncol(mx))
  rank.sum <- RankProd::RankProducts(
    data=mx,
    cl=cl.down,
    calculateProduct=FALSE,
    gene.names = row.names(ranks.down.mx.2)
    )
  rank.sum
}
```

```{r}
get.rank.sum.stats <- function(rank.sum.result) {
  # Get rank sums:
  RS <- as.data.frame(rank.sum.result$RSs) %>% 
    tibble::rownames_to_column(.) %>% 
    dplyr::select(1:2) %>% 
    set_colnames(c("Gene", "Rank.Sum")) %>% 
    arrange(Rank.Sum)
  
  # Get p values:
  PVAL <- as.data.frame(rank.sum.result$pval) %>% 
  tibble::rownames_to_column(.) %>% 
  dplyr::select(1:2) %>% 
  set_colnames(c("Gene", "P.Val"))
  
  # Merge:
  RS.PVAL <- merge(x=RS, y=PVAL, by="Gene", all=T)
  
  # Adjust p values:
  RS.PVAL$P.Val.Adj <- p.adjust(RS.PVAL$P.Val, method = "fdr")
  
  RS.PVAL
}
```


### Rank sum stats: downregulation  
```{r message=FALSE}
rank.sum.down <- get.rank.sum(ranks.down.mx.2) %>% 
  get.rank.sum.stats(.)
# class(rank.sum.down) # "data.frame"
# dim(rank.sum.down) # 11684     4
rank.sum.down %>% 
  arrange(Rank.Sum) %>% 
  head(., 10) %>% 
  knitr::kable(., row.names = F, align=rep("l", 4), format = "html", escape = F) %>% 
  kableExtra::kable_styling("striped", full_width = T)
```

### Rank sum stats: upregulation  
```{r message=FALSE}
rank.sum.up <- get.rank.sum(ranks.up.mx.2) %>% 
  get.rank.sum.stats(.)
# class(rank.sum.up) # "data.frame"
# dim(rank.sum.up) # 12786     4
rank.sum.up %>% 
  arrange(Rank.Sum) %>% 
  head(., 10) %>% 
  knitr::kable(., row.names = F, align=rep("l", 4), format = "html", escape = F) %>% 
  kableExtra::kable_styling("striped", full_width = T)
```

## Heatmaps for top regulated genes  
([Datanovia Tutorial](https://www.datanovia.com/en/lessons/heatmap-in-r-static-and-interactive-visualization/#r-base-heatmap-heatmap))

### Heatmap: 100 most downregulated genes  
```{r}
# nrow(subset(rank.sum.up, P.Val.Adj <= 0.05)) # 594
# nrow(subset(rank.sum.up, P.Val.Adj <= 0.01)) # 317
down.100 <- dplyr::arrange(rank.sum.down, P.Val.Adj, Rank.Sum) %>% 
  head(.,100) %>% .$Gene
# down.100
mx.down.100 <- ranks.down.mx.2[down.100,] %>% 
  set_colnames(names(all.data.list))
# dim(mx.down.100) # 100   5
head(mx.down.100)
```

```{r}
down.100.df <- as.data.frame(t(mx.down.100))
# dim(down.100.df) # 5 100

row.names(down.100.df) <- names(all.data.list)

down.100.df <- as.data.frame(t(mx.down.100)) %>% 
  set_rownames(names(all.data.list)) %>% 
  tibble::rownames_to_column("Dataset") %>% 
  dplyr::mutate(Technology = "Microarray") %>%
  dplyr::mutate(Cell.type = "Macrophages") %>%
  dplyr::select(Dataset, Technology, Cell.type, dplyr::everything())
down.100.df$Technology[down.100.df$Dataset %in% names(rna.seq.data.list)] <- "RNA-seq"
down.100.df$Cell.type[down.100.df$Dataset %in% c("GSE64179", "A-BUGS-23_DC")] <- "Dendritic cells"
down.100.df[,1:6]
```

```{r}
col <- list(Technology = c("Microarray" = "yellow", "RNA-seq" = "darkgreen"),
            Cell.type = c("Macrophages" = "lightgrey", "Dendritic cells" = "black"))
ha <- ComplexHeatmap::HeatmapAnnotation(
  Technology = down.100.df$Technology,
  Cell.type = down.100.df$Cell.type,
  col=col
)
```

```{r}
# help(Heatmap, package="ComplexHeatmap")
# "YlOrRd", "YlOrBr", "YlGnBu"*, "PuBuGn", "YlGn", "RdPu"-, "PuRd"-, "BuGn"
# "Blues", "Greens", "Purples", "Oranges"
heat.cols <- colorRampPalette(RColorBrewer::brewer.pal(9, "YlGnBu"))(256) %>% rev
ComplexHeatmap::Heatmap(
  mx.down.100, name = "Rank",
  col = heat.cols,
  top_annotation = ha)
```


### Heatmap: 100 most upregulated genes  
```{r}
up.100 <- dplyr::arrange(rank.sum.up, P.Val.Adj, Rank.Sum) %>% 
  head(.,100) %>% .$Gene
head(up.100)
mx.up.100 <- ranks.up.mx.2[up.100,] %>% 
  set_colnames(names(all.data.list))
 dim(mx.up.100) # 100   6
head(mx.up.100)
```


```{r}
up.100.df <- as.data.frame(t(mx.up.100))
# dim(down.100.df) # 5 100
row.names(up.100.df) <- names(all.data.list)

up.100.df <- as.data.frame(t(mx.up.100)) %>% 
  set_rownames(names(all.data.list)) %>% 
  tibble::rownames_to_column("Dataset") %>% 
  dplyr::mutate(Technology = "Microarray") %>%
  dplyr::mutate(Cell.type = "Macrophages") %>%
  dplyr::select(Dataset, Technology, Cell.type, dplyr::everything())
up.100.df$Technology[up.100.df$Dataset %in% names(rna.seq.data.list)] <- "RNA-seq"
up.100.df$Cell.type[up.100.df$Dataset %in% c("GSE64179", "A-BUGS-23_DC")] <- "Dendritic cells"
up.100.df[,1:6]
```

```{r}
col <- list(Technology = c("Microarray" = "yellow", "RNA-seq" = "darkgreen"),
            Cell.type = c("Macrophages" = "lightgrey", "Dendritic cells" = "black"))
ha <- ComplexHeatmap::HeatmapAnnotation(
  Technology = up.100.df$Technology,
  Cell.type = up.100.df$Cell.type,
  col=col
)
```

```{r}
# help(Heatmap, package="ComplexHeatmap")
# "YlOrRd", "YlOrBr", "YlGnBu"*, "PuBuGn", "YlGn", "RdPu"-, "PuRd"-, "BuGn"
# "Blues", "Greens", "Purples", "Oranges"
heat.cols <- colorRampPalette(RColorBrewer::brewer.pal(9, "YlGnBu"))(256) %>% rev
ComplexHeatmap::Heatmap(
  mx.up.100, name = "Rank",
  col = heat.cols,
  top_annotation = ha)
```



## Save genes  
### Save top downregulated genes  
```{r results="asis"}
gs.down.01 <- rank.sum.down %>% 
  dplyr::filter(P.Val.Adj <= 0.01) %>% 
  use_series("Gene")
 length(gs.down.01) # 371
cat("Number of consistently <b><i>down</i></b>regulated genes (p <= 0.01):<b>", length(gs.down.01), "</b>")
```

```{r}
out.file.down.01 <- paste0("data_out/", this.script, ".down.01.txt")
cat(gs.down.01, sep="\n", file=out.file.down.01)
cat("File saved:", out.file.down.01)
```


### Save top upregulated genes  
```{r results="asis"}
gs.up.01 <- rank.sum.up %>% 
  dplyr::filter(P.Val.Adj <= 0.01) %>% 
  use_series("Gene")
length(gs.up.01) # 547
cat("Number of consistently <b><i>up</i></b>regulated genes (p <= 0.01):<b>", length(gs.up.01), "</b>")
```

```{r}
out.file.up.01 <- paste0("data_out/", this.script, ".up.01.txt")
cat(gs.up.01, sep="\n", file=out.file.up.01)
cat("File saved:", out.file.up.01)
```


## Session info  
<button class="button" onclick="myFunction('DIV_5')">Show/hide session info</button>
<div id="DIV_5" class="div_default_hide">

```{r print_date_and_time}
Sys.time()
```

```{r print_session_info}
# sessionInfo()
devtools::session_info()
```
</div>
  
<script>
  var divsToHide = document.getElementsByClassName("div_default_hide");
for(var i = 0; i < divsToHide.length; i++)
{
  divsToHide[i].style.display = 'none';
}
</script>
