---
title: "Volcano plots of microarray data"
author: "A. Nohturfft"
date: "`r Sys.Date()`"
output:
  html_notebook:
    df_print: paged
    toc: yes
    highlight: tango
    number_sections: no
    fig_caption: yes
    theme: sandstone
    code_folding: hide
    toc_depth: 3
    toc_float: yes
---
  
<style>
h1 {background: darkblue;color: white;padding-left: 7px;}
h2 {color: darkblue;}
.code-folding-btn {display: none;}
</style>
  
<script>
  function show_span(id) {
    var x = document.getElementById(id);
    if (x.style.display === 'none') {
      x.style.display = 'inline';
    } else {
      x.style.display = 'none';
    }
  }
function myFunction(id) {
  var x = document.getElementById(id);
  if (x.style.display === 'none') {
    x.style.display = 'block';
  } else {
    x.style.display = 'none';
  }
}
</script>

## Objective  

Generate volcano plots of microarray data.  
Overview:  

1. Load fold change data generated with Geo2r  
2. Aggregate by gene symbol; where the same gene is represented by several probes, calculate the _**median**_ fold change  
3. combine all datasets into a single data frame  
4. Generate plots using ggplot() / facet_wrap()  

  
```{r include=FALSE}
# About <style> chunk above:
# * provides custom formatting for level-1 and level-2 headers
# * hides the 'code' buttons next to every code chunk in the html output; leaves only one 'code' button at the top of the html notebook.

# About the <script> chunk at the top and bottom of the file:
# * allows to insert <div>'s whose content visbility is toggled with a button.
# * I use this to show/hide te output of sessionInfo() at the end of the script (see below)
```


```{r include=FALSE}
# RMARKDOWN HELP
# Rmarkdown:
# https://bookdown.org/yihui/rmarkdown/

# Chunk options:
# https://yihui.name/knitr/options
```

```{r cleanup, warning=FALSE, message=FALSE, error=FALSE}
# CHECK R VERSION
# stopifnot(R.version.string == "R version 3.4.3 (2017-11-30)")

# CLEANUP
# Clear all variables:
rm(list=ls(all=TRUE))

# Unload current packages:
# if (!is.null(names(utils::sessionInfo()[["otherPkgs"]]))) pacman::p_unload("all")
```

```{r setup}
# KNITR DEFAULTS
knitr::opts_chunk$set(eval = TRUE, fig.height = 10)
```

```{r packages, results="hide", warning=FALSE, message=FALSE, error=FALSE}
# LOAD PACKAGES:
if (!require(magrittr)) install.packages("magrittr")
library(magrittr)
if (!require(dplyr)) install.packages("dplyr")
library(dplyr)
if (!require(scales)) install.packages("scales")
library(scales)
if (!require(knitr)) install.packages("knitr")
library(knitr)
if (!require(kableExtra)) install.packages("kableExtra")
library(kableExtra)
if (!require(ggplot2)) install.packages("ggplot2")
library(ggplot2)
```

```{r setup_script, eval=TRUE, results='hide'}
# GET CURRENT SCRIPT NAME
(this.script <- rstudioapi::getActiveDocumentContext() %>% .$path %>% basename)
getwd()
list.files()
stopifnot(this.script != "")
```

## GEO2R data  

### Geo2r data files  
```{r results="hide"}
geo2r.data.files <- list.files(path="../source_data", recursive = T, 
                               pattern="^GSE.*Geo2r.*txt$", full.names = T)
cat(basename(geo2r.data.files), sep="<br>\n")
```

```{r results="hide"}
tailleux.data.files <- list.files(path="../source_data", recursive = T,
                                  pattern="*tailleux.*txt", full.names = T)
cat(basename(tailleux.data.files), sep="<br>\n")

```

```{r results="asis"}
muarray.data.files <- c(geo2r.data.files, tailleux.data.files)
muarray.data.files <- grep("GSE29731", muarray.data.files, invert = T, value=T)
cat(basename(muarray.data.files), sep="<br>\n")
```

### Load Geo2r data  
```{r results="hide"}
geo2r.data.list <- lapply(muarray.data.files, function(f) {
  a <- read.table(f, sep="\t", header=T, quote='"', stringsAsFactors = F)
  names(a)[grep("symbol", names(a), ignore.case = T)] <- "Gene.symbol"
  names(a) <- stringr::str_remove(names(a), "^X\\.")
  names(a) <- stringr::str_remove(names(a), "\\.$")
  a
}) %>% 
  set_names(stringr::str_remove_all(basename(muarray.data.files), "\\.txt"))

names(geo2r.data.list) <- stringr::str_remove(names(geo2r.data.list), "^X")
names(geo2r.data.list)[grep("DC", names(geo2r.data.list))] <- "A-BUGS-23_DC"
names(geo2r.data.list)[grep("MP", names(geo2r.data.list))] <- "A-BUGS-23_MP"
names(geo2r.data.list) <- stringr:: str_remove(names(geo2r.data.list), "_Geo2r.*$")

# class(geo2r.data.list) # "list"
# length(geo2r.data.list) # 5
# sapply(geo2r.data.list, class)
# sapply(geo2r.data.list, nrow) %>% unname # 29102 72000 47231 24501 24501
```

```{r}
data.frame(List.item = names(geo2r.data.list),
           class = sapply(geo2r.data.list, class),
           Columns = sapply(geo2r.data.list, ncol),
           Rows = comma(sapply(geo2r.data.list, nrow)),
           Has.Gene.Symbols = sapply(geo2r.data.list, function(x) {any(grepl("Gene.symbol", names(x)))})
           # Unique.genes = sapply(geo2r.data.list, function(x) {length(unique(x$))})
           ) %>% 
  mutate(Has.Gene.Symbols = cell_spec(Has.Gene.Symbols, "html", color = ifelse(Has.Gene.Symbols == "TRUE", "black", "red"))) %>% 
  knitr::kable(., row.names = F, align=c("l", "l", "c", "c", "c"), format = "html", escape = F) %>% 
  kable_styling("striped", full_width = T)
```

```{r}
head(geo2r.data.list[[2]])
```


## Aggregate logFC by gene symbol  
```{r results="hide"}
# help(do, package="dplyr")
get.max.fc.row <- function(dafra) {
  dafra$tmp <- abs(dafra$logFC)
  the.max <- max(dafra$tmp)
  indx <- which(dafra$tmp==the.max)[1]
  dafra[indx,-which(names(dafra)=="tmp")]
}
all.df <- lapply(seq_along(geo2r.data.list), function(i, list.names) {
  df <- geo2r.data.list[[i]]
  x <- dplyr::select(df, Gene.symbol, logFC, P.Value) %>% 
    dplyr::mutate(Gene.symbol = ifelse(Gene.symbol=="", NA, Gene.symbol)) %>% 
    tidyr::drop_na(.) %>% 
    dplyr::mutate(logFC = as.numeric(logFC)) %>% 
    dplyr::mutate(Gene.symbol = stringr::str_remove_all(Gene.symbol, '\\"')) %>% 
    tidyr::drop_na(.) %>% 
    dplyr::group_by(Gene.symbol) %>% 
    dplyr::do(get.max.fc.row(.)) %>% 
    tidyr::drop_na(.) %>% 
    dplyr::mutate(Dataset = names(geo2r.data.list)[i]) %>% 
    dplyr::select(Dataset, dplyr::everything())
  x
}, list.names = names(geo2r.data.list)) %>% 
  do.call(rbind, .)

```

```{r}
# dim(all.df) # 69323     4
head(all.df)
```


## Plot  
```{r}
gg1 <- all.df %>% 
  ggplot(aes(x=logFC, y=-log10(P.Value))) +
  geom_point(alpha=0.4) +
  facet_wrap(~Dataset, scales="free") +
  theme(strip.text.x = element_text(size = 10, face="bold"))
gg1
```

Use a different color for significant genes:  
```{r}
d1 <- dplyr::filter(all.df, (abs(logFC) >= 1 & P.Value <= 0.01))
d2 <- dplyr::filter(all.df, (abs(logFC) < 1 | P.Value > 0.01))
```

```{r message=FALSE, warning=FALSE}
font.size <- 10
gg2 <- ggplot(data=d2, mapping=aes(x=logFC, y=-log10(P.Value))) +
  geom_point(colour="gray", alpha=0.2) +
  geom_point(data=d1, alpha=0.4, colour="blue") +
  facet_wrap(~Dataset, scales="free") +
  theme(strip.text.x = element_text(size = 12, face="bold"),
        text = element_text(size=font.size, face="bold"),
            axis.text.x = element_text(size=font.size, face="bold"),
            axis.text.y = element_text(size=font.size, face="bold"),
            axis.title.x = element_text(size=font.size, face="bold"),
            axis.title.y = element_text(size=font.size, face="bold"),
            plot.title = element_text(size=font.size, face="bold")) +
  scale_x_continuous(limits=c(-7,7))
gg2
```

Save plot as PDF:  
```{r}
out.file.2 <- paste0(this.script, ".plot.pdf")
ggsave(filename=out.file.2, plot = gg2, width = 16, height= 16, units="cm")
cat("Saved:", out.file.2)
```



## Session info  
<button class="button" onclick="myFunction('DIV_5')">Show/hide session info</button>
<div id="DIV_5" class="div_default_hide">

```{r print_date_and_time}
Sys.time()
```

```{r print_session_info}
# sessionInfo()
devtools::session_info()
```
</div>
  
<script>
  var divsToHide = document.getElementsByClassName("div_default_hide");
for(var i = 0; i < divsToHide.length; i++)
{
  divsToHide[i].style.display = 'none';
}
</script>
