---
title: "R Notebook"
author: '9401'
date: "14/07/2020"
output:
  html_notebook:
    toc: yes
    toc_depth: 3
    toc_float: yes
    highlight: tango
    number_sections: no
    fig_caption: yes
    theme: sandstone
    code_folding: hide
---

  
<style>
h1 {background: darkblue;color: white;padding-left: 7px;}
h2 {color: darkblue;}
.code-folding-btn {display: none;}
</style>
  
<script>
  function show_span(id) {
    var x = document.getElementById(id);
    if (x.style.display === 'none') {
      x.style.display = 'inline';
    } else {
      x.style.display = 'none';
    }
  }
function myFunction(id) {
  var x = document.getElementById(id);
  if (x.style.display === 'none') {
    x.style.display = 'block';
  } else {
    x.style.display = 'none';
  }
}
</script>

```{r cleanup, warning=FALSE, message=FALSE, error=FALSE}
# CHECK R VERSION
 stopifnot(R.version.string == "R version 3.5.3 (2019-03-11)")

# CLEANUP
# Clear all variables:
rm(list=ls(all=TRUE))

# Unload current packages:
if (!is.null(names(utils::sessionInfo()[["otherPkgs"]]))) pacman::p_unload("all") 
```

```{r setup}
# KNITR DEFAULTS
knitr::opts_chunk$set(eval = TRUE, fig.height = 10)
```


## Load Packages
```{r, warning=FALSE, message=FALSE, error=FALSE}
library(readxl)
library(knitr)
library(magrittr)
library(dplyr)
library(kableExtra)
library(rstudioapi)
```


```{r setup_script, eval=TRUE, results='hide'}

# GET CURRENT SCRIPT NAME
(this.script <- rstudioapi::getActiveDocumentContext() %>% .$path %>% basename)
getwd()
list.files()
stopifnot(this.script != "")
```


## Table of individual studies chosen for the meta-analysis
```{r}
# getwd()
# list.files()
data.file <- "../data.in/datasets.final.xlsx"
stopifnot(file.exists(data.file))
datasets.final<- read_excel(path = data.file )

head(datasets.final)
names(datasets.final)

table.1 <- as.data.frame(datasets.final, row.names = FALSE) %>%
  rename(c("Sample Counts" = "Sample\r\nCount", "Control Replicates" = "Replicates\r\nCtrl" ,"Mtb Replicates"= "Replicates.MTb", "Time point"= "Time point/hrs\r\nchosen")) 
  head(table.1)
  names(table.1)
  
table.1.sub <- subset(table.1 ) %>%
  dplyr::select("Series_ID","Pubmed ID","Sample Counts","Mtb Replicates", "Control Replicates","Time point", "Tissue","Treatment" , "Technology",  "Platform ID")

  knitr::kable(table.1.sub, row.names = FALSE, format = "html") %>%
  kableExtra::kable_styling("striped", full_width = T)
  
  
```

