---
title: "GSE29731: Add gene symbols to Geo2r output"
author: "A. Nohturfft"
date: "`r Sys.Date()`"
output:
  html_notebook:
    df_print: paged
    toc: yes
    highlight: tango
    number_sections: no
    fig_caption: yes
    theme: sandstone
    code_folding: hide
    toc_depth: 3
    toc_float: yes
---
  
<style>
h1 {background: darkblue;color: white;padding-left: 7px;}
h2 {color: darkblue;}
.code-folding-btn {display: none;}
</style>
  
<script>
  function show_span(id) {
    var x = document.getElementById(id);
    if (x.style.display === 'none') {
      x.style.display = 'inline';
    } else {
      x.style.display = 'none';
    }
  }
function myFunction(id) {
  var x = document.getElementById(id);
  if (x.style.display === 'none') {
    x.style.display = 'block';
  } else {
    x.style.display = 'none';
  }
}
</script>
  
```{r include=FALSE}
# About <style> chunk above:
# * provides custom formatting for level-1 and level-2 headers
# * hides the 'code' buttons next to every code chunk in the html output; leaves only one 'code' button at the top of the html notebook.

# About the <script> chunk at the top and bottom of the file:
# * allows to insert <div>'s whose content visbility is toggled with a button.
# * I use this to show/hide te output of sessionInfo() at the end of the script (see below)
```


```{r include=FALSE}
# RMARKDOWN HELP
# Rmarkdown:
# https://bookdown.org/yihui/rmarkdown/

# Chunk options:
# https://yihui.name/knitr/options
```

```{r cleanup, warning=FALSE, message=FALSE, error=FALSE}
# CHECK R VERSION
# stopifnot(R.version.string == "R version 3.4.3 (2017-11-30)")

# CLEANUP
# Clear all variables:
rm(list=ls(all=TRUE))

# Unload current packages:
# if (!is.null(names(utils::sessionInfo()[["otherPkgs"]]))) pacman::p_unload("all")
```

```{r setup}
# KNITR DEFAULTS
knitr::opts_chunk$set(eval = TRUE, fig.height = 10)
```

```{r packages, results="hide", warning=FALSE, message=FALSE, error=FALSE}
# LOAD PACKAGES:
library(magrittr)
library(dplyr)
library(tidyr)
library(biomaRt)
library(AnnotationDbi) 
# BiocManager::install("org.Mm.eg.db")
library(org.Mm.eg.db)
library(scales)
```

```{r setup_script, eval=TRUE, results='hide'}
# GET CURRENT SCRIPT NAME
(this.script <- rstudioapi::getActiveDocumentContext() %>% .$path %>% basename)
getwd()
list.files()
stopifnot(this.script != "")
```

Load data file
```{r}
data.file <- "GSE29731.txt"
stopifnot(file.exists(data.file))
df <- read.table(file = data.file, header = T, sep = "\t", stringsAsFactors = F)
# dim(df) # 57355     9
# cat(names(df), sep="\n")
# ID
# adj.P.Val
# P.Value
# t
# B
# logFC
# GB_ACC
# SEQUENCE
# SPOT_ID
head(df)
```

Do all probes have a proper gene accession number?
```{r results="asis"}
# 
cat("Are there any instances of a single space instead of a GB accesion? <b>",
    any(grepl(" +", df$GB_ACC)),
    "</b><br>\n<br>\n", sep="") # FALSE

cat("Are there any instances of EMPTY STRINGS instead of a GB accession? <b>",
    any(df$GB_ACC ==""),
    "</b><br>\n<br>\n", sep="") # FALSE
```

Removing probes that don't have a proper gene bank accession id ...  
```{r results="asis"}
df2 <- filter(df, GB_ACC !="") # 
cat("Original number of data rows:",  comma(nrow(df)), "<br>\n<br>\n")
cat("Number of data rows after filter:", comma(nrow(df2)))
```



```{r results="asis"}
cat("Are any Genbank accession IDs duplicated? <b>",
    any(duplicated(df2$GB_ACC)),
    "</b><br>\n<br>\n", sep="") # TRUE

cat("How many Genbank accession ID duplicates? <b>",
    comma(sum(duplicated(df2$GB_ACC))),
    "</b><br>\n<br>\n", sep="") # 46,369

gb.acc.unique <- unique(df2$GB_ACC) # leaving only probes with a unique gene bank accession id

cat("Number of unique Genbank accession IDs? <b>",
    comma(length(gb.acc.unique)),
    "</b>", sep="") # 25,631
```


## Mapping to gene symbols: Biomart  
Load Biomart data and generate a table that maps Genbank accession IDs to gene symbols.  
```{r}
head(sort(unique(df2$GB_ACC)))
```

```{r messsage=FALSE}
# head(df2$GB_ACC)
# mart = useMart('ensembl')
# ds <- listDatasets(mart)
# grep("musculus", ds$dataset, ignore.case=T, value=T)
ensembl = useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "mmusculus_gene_ensembl" )
# mouse.attr <- listAttributes(ensembl)
# names(mouse.attr) # "name"        "description" "page"
# grep("refseq", mouse.attr$name, ignore.case = T, value=T)
# grep("symbol", mouse.attr$name, ignore.case = T, value=T)
Mapped.ids <- getBM(attributes = c("refseq_mrna","mgi_symbol"),
        filters = "refseq_mrna",
        values = sort(unique(df2$GB_ACC)),
        mart = ensembl)
# dim(Mapped.ids) # 17804     2
# sum(complete.cases(Mapped.ids)) # 17804
# sapply(Mapped.ids, class)
print(head(Mapped.ids), row.names = F)
```

```{r results="asis"}
cat("Number of unique Genbank IDs in GEO2R data file: <b>",
    comma(length(unique(df2$GB_ACC))),
    "</b><br>\n<br>\n", sep="")
# 30,679

cat("Number of unique Genbank IDs in Biomart table: <b>",
    comma(length(unique(Mapped.ids$refseq_mrna))),
    "</b><br>\n<br>\n", sep="")
# 19,794

cat("Are all of the Genbank IDs in the GEO2R data file represented in the biomart table? <b>",
    all(df$GB_ACC %in% Mapped.ids$refseq_mrna),
    "</b><br>\n<br>\n", sep="")
# FALSE

percent(sum(gb.acc.unique %in% Mapped.ids$refseq_mrna) / length(gb.acc.unique)) %>% 
  cat("How many Genbanks IDs from the GEO2R file could be mapped with Biomart? <b>",
    comma(sum(df$GB_ACC %in% Mapped.ids$refseq_mrna)),
    " (", ., ")",
    "</b><br>\n", sep="")
# 24,685 (64.5%)
```

```{r}
gb.ids.not.mapped <- sort(unique(setdiff(df2$GB_ACC, Mapped.ids$refseq_mrna)))
cat("Number of Genbank IDs that could not be mapped with Biomart: <b>",
    comma(length(gb.ids.not.mapped)),
    "</b><br>\n", sep="")
```


Some genbank IDs that could not be mapped:
```{r}
cat(head(gb.ids.not.mapped), sep=", ")
```

```{r}
cat(tail(gb.ids.not.mapped), sep=", ")
```

## Mapping to gene symbols: Bioconductor genome annotations  
```{r message=FALSE}
AnnotationDbi::keytypes(org.Mm.eg.db)
Mapped.ids.2 <- AnnotationDbi::select(org.Mm.eg.db,
                       keys = unique(df2$GB_ACC),
                       columns = c("REFSEQ","SYMBOL"),
                       keytype = "REFSEQ") 
Mapped.ids.3 <- Mapped.ids.2 %>%
  filter( !is.na(SYMBOL))
# dim(Mapped.ids.3) # 2976    2
# all(df2$GB_ACC %in% Mapped.ids.3$REFSEQ) # FALSE
# length(unique(df2$GB_ACC)) # 25,631
sum(unique(df2$GB_ACC) %in% Mapped.ids.3$REFSEQ) # 18,700
head(Mapped.ids.3)
```

Combine the Biomart and Bioconductor mapping tables with `rbind`:  
```{r}
# When combining two data frames with rbind, the column names must be the same:
# names(Mapped.ids) # "refseq_mrna" "hgnc_symbol"
# names(Mapped.ids.3) # "REFSEQ" "SYMBOL"
names(Mapped.ids) <- names(Mapped.ids.3)
mapped.ids.all <- rbind(Mapped.ids, Mapped.ids.3) %>% 
  dplyr::distinct()
head(mapped.ids.all)
```

```{r}
percent(sum(gb.acc.unique %in% mapped.ids.all$REFSEQ) / length(gb.acc.unique)) %>% 
  cat("How many Genbanks IDs from the GEO2R file could be mapped now? <b>",
    comma(sum(df2$GB_ACC %in% mapped.ids.all$REFSEQ)),
    " (", ., ")",
    "</b><br>\n", sep="")
```

Join GEO2R data and mapping table to add gene symbols:  
```{r}
df3 <- merge(x=df2, y=mapped.ids.all, by.x="GB_ACC", by.y="REFSEQ", all.x=F, all.y=F) %>% 
  dplyr::rename(Gene.symbol = SYMBOL) %>% 
  dplyr::select(ID, adj.P.Val, P.Value, t, B, logFC, Gene.symbol)
nrow(df)  # 72,000
nrow(df2) # 72,000
nrow(df3) # 56,868
```

## Map mouse symbols to human gene symbols  
### Load mouse/human symbol mapping data
```{r}
hum.ms.map.file <- "../../mouse_human_symbols_2018_03_05b.txt"
stopifnot(file.exists(hum.ms.map.file))
ms.hu.map.df <- read.table(file=hum.ms.map.file, sep="\t", header=T,comment.char = "#",
                           stringsAsFactors = F)
head(ms.hu.map.df)
```

### Merge/map  
```{r}
all(df3$Gene.symbol %in% ms.hu.map.df$mouse) # FALSE
comma(nrow(df3)) # 56,868
sum(df3$Gene.symbol %in% ms.hu.map.df$mouse) # 49933
chunk.match <- subset(df3, Gene.symbol %in% ms.hu.map.df$mouse)
chunk.no.match <- subset(df3, !Gene.symbol %in% ms.hu.map.df$mouse) %>% 
  dplyr::mutate(Gene.symbol = toupper(Gene.symbol))
chunk.match.merge <- merge(x=chunk.match, y=ms.hu.map.df,
                           by.x="Gene.symbol", by.y="mouse",
                           all.x=T, all.y=F) %>% 
  dplyr::select(-Gene.symbol) %>% 
  dplyr::rename(Gene.symbol = human)
head(chunk.match.merge)
```

```{r}
df4 <- rbind(chunk.no.match, chunk.match.merge) %>% 
  dplyr::arrange(Gene.symbol)
dim(df3) # 56,868     7
dim(df4) # 64,737     7
```


## Save results:  
```{r}
out.file <- "GSE29731_Geo2r_symbols.txt"
cat("#", this.script, "\n", file=out.file, append=F)
cat("#", Sys.time(), "\n", file=out.file, append=T)
cat("# Input file:", data.file, "\n", file=out.file, append=T)
cat("# Author: A. Nohturfft", "\n", file=out.file, append=T)
write.table(df4, file=out.file, row.names = F, col.names = T, sep="\t", append=T)
cat("Saved:", out.file)
```


## Session info  
<button class="button" onclick="myFunction('DIV_5')">Show/hide session info</button>
<div id="DIV_5" class="div_default_hide">

```{r print_date_and_time}
Sys.time()
```

```{r print_session_info}
# sessionInfo()
devtools::session_info()
```
</div>
  
<script>
  var divsToHide = document.getElementsByClassName("div_default_hide");
for(var i = 0; i < divsToHide.length; i++)
{
  divsToHide[i].style.display = 'none';
}
</script>
