---
title: "Initial exploration of GSE148731 data"
author: "A. Nohturfft"
date: "`r Sys.Date()`"
output:
  html_notebook:
    df_print: paged
    toc: yes
    highlight: tango
    number_sections: no
    fig_caption: yes
    theme: sandstone
    code_folding: hide
    toc_depth: 3
    toc_float: yes
---
  
<style>
h1 {background: darkblue;color: white;padding-left: 7px;}
h2 {color: darkblue;}
.code-folding-btn {display: none;}
</style>
  
<script>
  function show_span(id) {
    var x = document.getElementById(id);
    if (x.style.display === 'none') {
      x.style.display = 'inline';
    } else {
      x.style.display = 'none';
    }
  }
function myFunction(id) {
  var x = document.getElementById(id);
  if (x.style.display === 'none') {
    x.style.display = 'block';
  } else {
    x.style.display = 'none';
  }
}
</script>

## Objectives  

1. Analyze processed fold change data for GSE148731 available on GEO (24 hr xlsx files)  
2. Plot distriubtion of fold changes  
3. Generate volcano plot  
4. Save FC and p.values as text file for subsequent meta analysis  
  
```{r include=FALSE}
# About <style> chunk above:
# * provides custom formatting for level-1 and level-2 headers
# * hides the 'code' buttons next to every code chunk in the html output; leaves only one 'code' button at the top of the html notebook.

# About the <script> chunk at the top and bottom of the file:
# * allows to insert <div>'s whose content visbility is toggled with a button.
# * I use this to show/hide te output of sessionInfo() at the end of the script (see below)
```


```{r include=FALSE}
# RMARKDOWN HELP
# Rmarkdown:
# https://bookdown.org/yihui/rmarkdown/

# Chunk options:
# https://yihui.name/knitr/options
```

```{r cleanup, warning=FALSE, message=FALSE, error=FALSE}
# CHECK R VERSION
# stopifnot(R.version.string == "R version 3.4.3 (2017-11-30)")

# CLEANUP
# Clear all variables:
rm(list=ls(all=TRUE))

# Unload current packages:
# if (!is.null(names(utils::sessionInfo()[["otherPkgs"]]))) pacman::p_unload("all")
```

```{r setup}
# KNITR DEFAULTS
knitr::opts_chunk$set(eval = TRUE, fig.height = 10)
```

```{r packages, results="hide", warning=FALSE, message=FALSE, error=FALSE}
# LOAD PACKAGES:
if (!requireNamespace("magrittr", quietly = TRUE)) install.packages("magrittr")
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
if (!requireNamespace("scales", quietly = TRUE)) install.packages("scales")
if (!requireNamespace("readxl", quietly = TRUE)) install.packages("readxl")
if (!requireNamespace("stringr", quietly = TRUE)) install.packages("stringr")
if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
library(magrittr)
library(dplyr)
library(scales)
library(readxl)
library(ggplot2)
```

```{r setup_script, eval=TRUE, results='hide'}
# GET CURRENT SCRIPT NAME
(this.script <- rstudioapi::getActiveDocumentContext() %>% .$path %>% basename)
getwd()
list.files()
stopifnot(this.script != "")
```

## Data files  
```{r database_file}
data.files <- list.files(pattern = "GSE148731.*xlsx", full.names = T)
stopifnot(all(file.exists(data.files)))
cat(basename(data.files), sep="\n")
```

## Load data  
```{r results="hide"}
df.list <- lapply(data.files, function(fil) {
  read_xlsx(path=fil,col_names = T) %>% 
    extract(,c("gene", "MEAS/CTRL or -CTRL/MEAS (scaled)", "p-value", "p-value (adjusted)")) %>% 
    set_colnames(c("Gene.symbol", "change", "P.Val", "Adj.P.Val"))
}) %>% 
  set_names(stringr::str_extract(basename(data.files), "MF[12]"))
# names(df.list) # "MF1" "MF2"
head(df.list[[1]])
```

```{r}
sapply(df.list, nrow) # 60450 60450 
sapply(df.list, function(df) {length(unique(df$Gene.symbol))}) # 58585 58585 
sapply(df.list, function(df) {min(df$change, na.rm = T)}) # -35.97831 -230.21264
sapply(df.list, function(df) {max(df$change, na.rm = T)}) # 4315.174 4781.793 
```


```{r}
df.all <- lapply(seq_along(df.list), function(i) {
  x <- df.list[[i]] %>% 
    mutate(Dataset=names(df.list[i])) %>% 
    select(Dataset, everything())
}) %>% 
  do.call(rbind, .)
dim(df.all) # 120900      5
```

```{r}
sum(df.list$MF1$change == 0, na.rm = T) # 20352
df.mf1 <- df.list$MF1 %>% 
  select(Gene.symbol, change) %>% 
  arrange(desc(change)) %>% 
  filter(complete.cases(.)) %>% 
  mutate(change=log(change)) %>% 
  mutate(num = seq_along(change))
# head(df.mf1)
plot(x=df.mf1$num, y=df.mf1$change)
df.mf2 <- df.list$MF2 %>% 
  select(Gene.symbol, change) %>% 
  arrange(change)
gg1 <- ggplot(data=df.mf1, aes(x=Gene.symbol, y=log(change))) +
  geom_point(size=1, color='blue') +
  # geom_point(data=df.mf2, aes(x=Gene.symbol, y=change))
  xlab(NULL)
gg1
```


```{r}
archs4.info <- h5read(archs4.db.file, "info")
as.vector(archs4.info) %>% as.matrix
# author        "Alexander Lachmann"                                   
# contact       "alexander.lachmann@mssm.edu"                          
# creation-date "2020-02-06"                                           
# lab           "Ma'ayan Lab - Icahn School of Medicine at Mount Sinai"
# version       "8"  
```


## Explore GEO series field  
```{r results="asis"}
archs4.series <- h5read(archs4.db.file, "meta/Sample_series_id")
cat("class(archs4.series):", class(archs4.series), "<br>\n")   # array
cat("length(archs4.series):", length(archs4.series), "<br>\n") # 238522
cat("dim(archs4.series):", dim(archs4.series), "<br>\n") # 238522
cat("head:", paste(head(archs4.series), collapse=","), "<br>\n")
cat("Number of unique series IDs:", comma(length(unique(archs4.series))), "<br>\n") # 7,909
```

## Search for GEO series IDs: exact matches  
```{r}
sum(archs4.series == "GSE64179")  # 0   GPL16791 	Illumina HiSeq 2500 (Homo sapiens)
sum(archs4.series == "GSE67427")  # 155 GPL16791 	Illumina HiSeq 2500 (Homo sapiens)
sum(archs4.series == "GSE114371") # 0   GPL17303 	Ion Torrent Proton (Homo sapiens)
sum(archs4.series == "GSE116405") # 0   GPL16791 	Illumina HiSeq 2500 (Homo sapiens)
sum(archs4.series == "GSE116412") # 0   GPL16791 	Illumina HiSeq 2500 (Homo sapiens)
sum(archs4.series == "GSE148731") # 0   GPL16791 	Illumina HiSeq 2500 (Homo sapiens)
```

## Search for GEO series IDs: PARTIAL matches  
### GSE64179  
Expected samples (based on GEO web page): 12  
```{r GSE64179, results="asis"}
sum(grepl("GSE64179", archs4.series)) %>% 
  cat("Number of matches found:", ., "<br>\n")
unique(grep("GSE64179", archs4.series, ignore.case=T, value=T)) %>% # "GSE64179Xx-xXGSE64183"
  cat("Unique match(es):", ., "<br>\n")
# Number of matches found: 12 <br>
# Unique match(es): GSE64179Xx-xXGSE64183 <br>
```

### GSE67427  
Expected samples (based on GEO web page): 156  
```{r GSE67427, results="asis"}
sum(grepl("GSE67427", archs4.series)) %>% 
  cat("Number of matches found:", ., "<br>\n")
unique(grep("GSE67427", archs4.series, ignore.case=T, value=T)) %>% # 
  cat("Unique match(es):", ., "<br>\n")
# Expected samples (based on GEO web page): 156 <br>
# Number of matches found: 155 <br>
# Unique match(es): GSE67427 <br>
```

### GSE114371  
Platform: GPL17303, Ion Torrent Proton (Homo sapiens)  
Expected samples (based on GEO web page): 36  
```{r GSE114371, results="asis"}
sum(grepl("GSE114371", archs4.series)) %>% 
  cat("Number of matches found:", ., "<br>\n")
unique(grep("GSE114371", archs4.series, ignore.case=T, value=T)) %>% # 
  cat("Unique match(es):", ., "<br>\n")
```

### GSE116412  
GEO:  
This SuperSeries is composed of the following SubSeries:  

* GSE116399 	Gene activation precedes DNA demethylation in response to infection in human dendritic cells  
* GSE116405 	Gene activation precedes DNA demethylation in response to infection in human dendritic cells  
* GSE116406 	Gene activation precedes DNA demethylation in response to infection in human dendritic cells  

### GSE116399  
Public on Jul 03, 2018  
Platform: GPL16791, Illumina HiSeq 2500 (Homo sapiens)  
Expected samples (based on GEO web page): 53  
```{r GSE116399, results="asis"}
sum(grepl("GSE116399", archs4.series)) %>% 
  cat("Number of matches found:", ., "<br>\n")
unique(grep("GSE116399", archs4.series, ignore.case=T, value=T)) %>% # 
  cat("Unique match(es):", ., "<br>\n")
```

### GSE116405  
Public on Jul 03, 2018  
Platform: GPL16791, Illumina HiSeq 2500 (Homo sapiens)  
Expected samples (based on GEO web page): 51  
Google search (4-Jul-20): 'Your search - "archs4" "GSE116405" - did not match any documents.'  
```{r GSE116405, results="asis"}
sum(grepl("GSE116405", archs4.series)) %>% 
  cat("Number of matches found:", ., "<br>\n")
unique(grep("GSE116405", archs4.series, ignore.case=T, value=T)) %>% # 
  cat("Unique match(es):", ., "<br>\n")
```

### GSE116406  
Public on Jul 03, 2018  
Submission date: Jun 28, 2018  
Last update date: Apr 17, 2019  
Platform: GPL16791, Illumina HiSeq 2500 (Homo sapiens)  
Expected samples (based on GEO web page): 10  
```{r GSE116406, results="asis"}
sum(grepl("GSE116406", archs4.series)) %>% 
  cat("Number of matches found:", ., "<br>\n")
unique(grep("GSE116406", archs4.series, ignore.case=T, value=T)) %>% # 
  cat("Unique match(es):", ., "<br>\n")
```

### GSE148731  
Public on Apr 16, 2020  
Submission date: Apr 15, 2020  
Last update date: May 07, 2020  
Platform: GPL16791, Illumina HiSeq 2500 (Homo sapiens)  
Expected samples (based on GEO web page): 24  
```{r GSE148731, results="asis"}
sum(grepl("GSE148731", archs4.series)) %>% 
  cat("Number of matches found:", ., "<br>\n")
unique(grep("GSE148731", archs4.series, ignore.case=T, value=T)) %>% # 
  cat("Unique match(es):", ., "<br>\n")
```

## GEO sample (GSM) IDs  
```{r}
archs4.samples = h5read(archs4.db.file, "meta/Sample_geo_accession")
# comma(length(samples)) # 238,522
# sample.num <- stringr::str_remove(samples, "GSM") %>% as.numeric %>% sort(., decreasing = T)
# class(sample.num)
# head(sample.num)
# 4230225 4230224 4230223 4230222 4230221 4230220

# GSM4230225 >> Series GSE142479
# Public on Feb 14, 2020
```



## Session info  
<button class="button" onclick="myFunction('DIV_5')">Show/hide session info</button>
<div id="DIV_5" class="div_default_hide">

```{r print_date_and_time}
Sys.time()
```

```{r print_session_info}
# sessionInfo()
devtools::session_info()
```
</div>
  
<script>
  var divsToHide = document.getElementsByClassName("div_default_hide");
for(var i = 0; i < divsToHide.length; i++)
{
  divsToHide[i].style.display = 'none';
}
</script>
