---
title: "Multidimensional scaling analysis"
author: "A. Nohturfft"
date: "`r Sys.Date()`"
output:
  html_notebook:
    df_print: paged
    toc: yes
    highlight: tango
    number_sections: no
    fig_caption: yes
    theme: sandstone
    code_folding: hide
    toc_depth: 3
    toc_float: yes
---
  
<style>
h1 {background: darkblue;color: white;padding-left: 7px;}
h2 {color: darkblue;}
.code-folding-btn {display: none;}
</style>
  
<script>
  function show_span(id) {
    var x = document.getElementById(id);
    if (x.style.display === 'none') {
      x.style.display = 'inline';
    } else {
      x.style.display = 'none';
    }
  }
function myFunction(id) {
  var x = document.getElementById(id);
  if (x.style.display === 'none') {
    x.style.display = 'block';
  } else {
    x.style.display = 'none';
  }
}
</script>

## Objective  

Perform multidimensional scaling analysis of all data, (including von Both).  
  
```{r include=FALSE}
# About <style> chunk above:
# * provides custom formatting for level-1 and level-2 headers
# * hides the 'code' buttons next to every code chunk in the html output; leaves only one 'code' button at the top of the html notebook.

# About the <script> chunk at the top and bottom of the file:
# * allows to insert <div>'s whose content visbility is toggled with a button.
# * I use this to show/hide te output of sessionInfo() at the end of the script (see below)
```


```{r include=FALSE}
# RMARKDOWN HELP
# Rmarkdown:
# https://bookdown.org/yihui/rmarkdown/

# Chunk options:
# https://yihui.name/knitr/options
```

```{r cleanup, warning=FALSE, message=FALSE, error=FALSE}
# CHECK R VERSION
# stopifnot(R.version.string == "R version 3.4.3 (2017-11-30)")

# CLEANUP
# Clear all variables:
rm(list=ls(all=TRUE))

# Unload current packages:
# if (!is.null(names(utils::sessionInfo()[["otherPkgs"]]))) pacman::p_unload("all")
```

```{r setup}
# KNITR DEFAULTS
knitr::opts_chunk$set(eval = TRUE, fig.height = 10)
```

```{r packages, results="hide", warning=FALSE, message=FALSE, error=FALSE}
# LOAD PACKAGES:
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!requireNamespace("enrichR", quietly = TRUE)) install.packages("enrichR")
if (!requireNamespace("ggrepel", quietly = TRUE)) install.packages("ggrepel")
if (!requireNamespace("ComplexHeatmap", quietly = TRUE)) BiocManager::install("ComplexHeatmap")
if (!require(stringr)) install.packages("stringr")
if (!require(rstudioapi)) install.packages("rstudioapi")
if (!require(tidyr)) install.packages("tidyr")
if (!require(tibble)) install.packages("tibble")
if (!require(readxl)) install.packages("readxl")
if (!require(purrr)) install.packages("purrr")
if (!require(RColorBrewer)) install.packages("RColorBrewer")
if (!require(devtools)) install.packages("devtools")
if (!require(magrittr)) install.packages("magrittr")
if (!require(dplyr)) install.packages("dplyr")
if (!require(scales)) install.packages("scales")
if (!require(knitr)) install.packages("knitr")
if (!require(kableExtra)) install.packages("kableExtra")
if (!requireNamespace("RankProd", quietly = TRUE)) BiocManager::install("RankProd")

library(magrittr)
library(dplyr)
library(scales)
library(knitr)
library(kableExtra)
library(RankProd)
library(ggrepel)
```

```{r setup_script, eval=TRUE, results='hide'}
# GET CURRENT SCRIPT NAME
(this.script <- rstudioapi::getActiveDocumentContext() %>% .$path %>% basename)
getwd()
list.files()
stopifnot(this.script != "")
```

## Data: microarray studies    

### Geo2r data files  
```{r results="hide"}
geo2r.data.files <- list.files(path="../source_data", recursive = T, 
                               pattern="^GSE.*Geo2r.*txt$", full.names = T)
cat(basename(geo2r.data.files), sep="<br>\n")
```

```{r results="hide"}
tailleux.data.files <- list.files(path="../source_data", recursive = T,
                                  pattern="*tailleux.*txt", full.names = T)
cat(basename(tailleux.data.files), sep="<br>\n")

```

```{r results="asis"}
muarray.data.files <- c(geo2r.data.files, tailleux.data.files)
muarray.data.files <- grep("GSE29731", muarray.data.files, invert = T, value=T)
cat(basename(muarray.data.files), sep="<br>\n")
```

### Load Geo2r data  
```{r results="hide"}
geo2r.data.list <- lapply(muarray.data.files, function(f) {
  a <- read.table(f, sep="\t", header=T, quote='"', stringsAsFactors = F)
  names(a)[grep("symbol", names(a), ignore.case = T)] <- "Gene.symbol"
  names(a) <- stringr::str_remove(names(a), "^X\\.")
  names(a) <- stringr::str_remove(names(a), "\\.$")
  a
}) %>% 
  set_names(stringr::str_remove_all(basename(muarray.data.files), "\\.txt"))

names(geo2r.data.list) <- stringr::str_remove(names(geo2r.data.list), "_Geo2r")
names(geo2r.data.list) <- stringr::str_replace(names(geo2r.data.list), "discovery", "Dis")
names(geo2r.data.list) <- stringr::str_replace(names(geo2r.data.list), "validation", "Val")
names(geo2r.data.list) <- stringr::str_remove(names(geo2r.data.list), "^X")
names(geo2r.data.list)[grep("DC", names(geo2r.data.list))] <- "A-BUGS-23_DC"
names(geo2r.data.list)[grep("MP", names(geo2r.data.list))] <- "A-BUGS-23_MP"
names(geo2r.data.list)[grep("GSE34151", names(geo2r.data.list))] <- "GSE34151_DC"
names(geo2r.data.list)[grep("GSE103092", names(geo2r.data.list))] <- "GSE103092_THP1"

# class(geo2r.data.list) # "list"
# length(geo2r.data.list) # 4
# sapply(geo2r.data.list, class) # "data.frame"
# sapply(geo2r.data.list, nrow) %>% unname # 29102 47231 24501 24501
```

```{r}
data.frame(List.item = names(geo2r.data.list),
           class = sapply(geo2r.data.list, class),
           Columns = sapply(geo2r.data.list, ncol),
           Rows = comma(sapply(geo2r.data.list, nrow)),
           Has.Gene.Symbols = sapply(geo2r.data.list, function(x) {any(grepl("Gene.symbol", names(x)))})
           # Unique.genes = sapply(geo2r.data.list, function(x) {length(unique(x$))})
           ) %>% 
  mutate(Has.Gene.Symbols = cell_spec(Has.Gene.Symbols, "html", color = ifelse(Has.Gene.Symbols == "TRUE", "black", "red"))) %>% 
  knitr::kable(., row.names = F, align=c("l", "l", "c", "c", "c"), format = "html", escape = F) %>% 
  kableExtra::kable_styling("striped", full_width = T)
```

### Aggregate logFC by gene symbol  
```{r}
geo.fold.change.list <- lapply(seq_along(geo2r.data.list), function(i, list.names) {
  df <- geo2r.data.list[[i]]
  x <- dplyr::select(df, Gene.symbol, logFC) %>% 
    dplyr::mutate(Gene.symbol = ifelse(Gene.symbol=="", NA, Gene.symbol)) %>% 
    tidyr::drop_na(.) %>% 
    dplyr::mutate(logFC = as.numeric(logFC)) %>% 
    dplyr::mutate(Gene.symbol = stringr::str_remove_all(Gene.symbol, '\\"')) %>% 
    tidyr::drop_na(.) %>% 
    dplyr::group_by(Gene.symbol) %>% 
    dplyr::summarise(logFC = median(logFC), .groups="drop") %>% 
    tidyr::drop_na(.) %>% 
    as.data.frame(., stringsAsFactors=FALSE)
  # names(x)[names(x) == "logFC"] <- list.names[i]
  x
}, list.names = names(geo2r.data.list)) %>% 
  set_names(names(geo2r.data.list))
names(geo.fold.change.list) <- stringr::str_remove(names(geo.fold.change.list), "_Geo2r.*$")
```


```{r}
data.frame(
  List.item = names(geo.fold.change.list),
  Class = sapply(geo.fold.change.list, function(x) {paste(class(x), collapse=",")}),
  Columns = sapply(geo.fold.change.list, ncol),
  Rows = comma(sapply(geo.fold.change.list, nrow)),
  Duplicated.genes = sapply(geo.fold.change.list, function(x) {any(duplicated(x$Gene.symbol))})
) %>% 
  knitr::kable(., row.names = F, align=c("l", "l", "c", "c", "c"), format = "html", escape = F) %>% 
  kableExtra::kable_styling("striped", full_width = T)
```

## Data: RNA-seq studies  
```{r}
# <a href="https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE64179">GSE64179</a>
# https://pubmed.ncbi.nlm.nih.gov/26392366/
add.geo.hyperlink <- function(id) {
  a <- "<a href='https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc="
  b <- "'>"
  c <- "</a>"
  res <- paste0(a, id, b, id, c)
  res
}

add.pubmed.hyperlink <- function(id) {
  a <- "<a href='https://pubmed.ncbi.nlm.nih.gov/"
  b <- "/'>"
  c <- "</a>"
  res <- paste0(a, id, b, id, c)
  res
}
```


```{r}
rs.dataset.table <- data.frame(
  c=c("GSE64179", "26392366", "6", "Dendritic cells"),
  b=c("GSE67427", "26586179", "8", "Monocyte derived macrophages"),
  a=c("GSE148731", "32341411", "6", "M1, M2 macrophages")
) %>%
  t %>% 
  as.data.frame(., stringsAsFactors=F) %>% 
  set_colnames(c("Accession", "Pubmed", "Replicates", "Cell.Type")) %>% 
  mutate(Accession=add.geo.hyperlink(Accession)) %>% 
  mutate(Pubmed = add.pubmed.hyperlink(Pubmed))
knitr::kable(rs.dataset.table, row.names = F, escape=F, format = "html") %>% 
  kableExtra::kable_styling("striped", full_width = T)
```


### Load RNA-seq data: GSE64179  
```{r message=FALSE}
# list.files("../source_data/rna_seq/GSE64179")
rna.seq.data.files <- list()
rna.seq.data.files$GSE64179_DC <- "../source_data/rna_seq/GSE64179/Pacis_2015_TableS6_DEG.xlsx"
stopifnot(file.exists(rna.seq.data.files$GSE64179_DC))
rna.seq.data.list <- list()
rna.seq.data.list$GSE64179_DC <- readxl::read_xlsx(path=rna.seq.data.files$GSE64179_DC, range = "B3:C18818") %>% 
  set_colnames(c("Gene.symbol", "logFC")) %>% 
  mutate(Gene.symbol = toupper(Gene.symbol)) %>% 
  group_by(Gene.symbol) %>% 
  summarise(logFC = median(logFC), .groups="drop") %>% 
  as.data.frame(., stringsAsFactors=FALSE)
```


```{r}
# WITHOUT GROUPING / AGGREGATING:
# dim(rna.seq.data$GSE64179) # 18815     2
# sapply(rna.seq.data$GSE64179, class) # "character"   "numeric"
# range(rna.seq.data$GSE64179$logFC) # -9.426613 11.612170
# sum(rna.seq.data$GSE64179$logFC == 0) # 0
# sum(duplicated(rna.seq.data$GSE64179$Gene.symbol)) # 37
# dups <- rna.seq.data$GSE64179$Gene.symbol[duplicated(rna.seq.data$GSE64179$Gene.symbol)]
# subset(rna.seq.data$GSE64179, Gene.symbol %in% dups) %>% 
#   arrange(Gene.symbol, logFC)
# head(rna.seq.data$GSE64179)
```

```{r}
# WITH AGGREGATING: 
# dim(rna.seq.data$GSE64179) # 18776     2
# sapply(rna.seq.data$GSE64179, class) # "character"   "numeric"
# range(rna.seq.data$GSE64179$logFC) # -9.426613 11.612170
# sum(rna.seq.data$GSE64179$logFC == 0) # 0
# sum(duplicated(rna.seq.data$GSE64179$Gene.symbol)) # 0
head(rna.seq.data.list$GSE64179_DC) %>% 
  knitr::kable(., row.names = F, format="html") %>% 
  kableExtra::kable_styling("striped", full_width = T)
```




### Load RNA-seq data: GSE67427
```{r}
rna.seq.data.files$GSE67427_MP <-"../source_data/rna_seq/GSE67427/GSE67427_table_s2.txt"
stopifnot(file.exists(rna.seq.data.files$GSE67427_MP))
#rna.seq.data.list <-list()
rna.seq.data.list$GSE67427_MP<- read.table(file = rna.seq.data.files$GSE67427_MP, header = T, sep = "\t", stringsAsFactors = FALSE)  %>% 
#str(rna.seq.data.list$GSE67427) #data.frame':	12728 obs. of  140 variables:
  # head(rna.seq.data.list$GSE67427)
  
  # names(rna.seq.data.list$GSE67427)
  # select(matches(c("name", "Rv.18")))
 select(matches(c("name","Rv.18.logFC"), ignore.case = TRUE)) %>%
   # head(rna.seq.data.list$GSE67427)
  set_colnames(c("Gene.symbol", "logFC")) %>%
  mutate(Gene.symbol = toupper(Gene.symbol)) %>% 
  group_by(Gene.symbol) %>% 
  summarise(logFC = median(logFC), .groups="drop") %>% 
  as.data.frame(., stringsAsFactors=FALSE)
 head(rna.seq.data.list$GSE67427_MP)
```

```{r}
# WITHOUT GROUPING / AGGREGATING:
dim(rna.seq.data.list$GSE67427_MP) #  12724     2
sapply(rna.seq.data.list$GSE67427_MP, class) # "character"   "numeric"
range(rna.seq.data.list$GSE67427_MP$logFC) # -4.155479  8.205439
sum(rna.seq.data.list$GSE67427_MP$logFC == 0) # 0
sum(duplicated(rna.seq.data.list$GSE67427_MP$Gene.symbol)) # 0
# dups <- rna.seq.data$GSE64179$Gene.symbol[duplicated(rna.seq.data$GSE64179$Gene.symbol)]
# subset(rna.seq.data$GSE64179, Gene.symbol %in% dups) %>% 
#   arrange(Gene.symbol, logFC)
# head(rna.seq.data$GSE64179)
```

```{r}
# WITH AGGREGATING: 
# dim(rna.seq.data$GSE64179) # 18776     2
# sapply(rna.seq.data$GSE64179, class) # "character"   "numeric"
# range(rna.seq.data$GSE64179$logFC) # -9.426613 11.612170
# sum(rna.seq.data$GSE64179$logFC == 0) # 0
# sum(duplicated(rna.seq.data$GSE64179$Gene.symbol)) # 0
head(rna.seq.data.list$GSE67427_MP) %>% 
  knitr::kable(., row.names = F, format="html") %>% 
  kableExtra::kable_styling("striped", full_width = T)

```

### Load RNA-seq data: GSE48731.MF1
```{r message=FALSE}
# list.files("../source_data/rna_seq/GSE148731")
# rna.seq.data.files <- list()
rna.seq.data.files$GSE148731.MF1 <- "../source_data/rna_seq/GSE148731/GSE148731_toptable_MF1_01.Rmd.2020_07_09.txt"
stopifnot(file.exists(rna.seq.data.files$GSE148731.MF1))
 # rna.seq.data.list <- list()
rna.seq.data.list$GSE148731.MF1 <- read.table(file = rna.seq.data.files$GSE148731.MF1, header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>%
   select(c("Gene.symbol", "logFC")) %>% 
   mutate(Gene.symbol = toupper(Gene.symbol)) %>% 
  # summarise(logFC = median(logFC), .groups="drop") %>% 
   as.data.frame(., stringsAsFactors=FALSE)
head(rna.seq.data.list$GSE148731.MF1)
```


```{r}
# WITH AGGREGATING:
dim(rna.seq.data.list$GSE148731.MF1) # 17802     2
sapply(rna.seq.data.list$GSE148731.MF1, class) # "character"   "numeric"
range(rna.seq.data.list$GSE148731.MF1$logFC) # -2.386124  9.704865
sum(rna.seq.data.list$GSE148731.MF1$logFC == 0) #
sum(duplicated(rna.seq.data.list$GSE148731.MF1$Gene.symbol)) # 0

```

```{r}
head(rna.seq.data.list$GSE148731.MF1) %>% 
  knitr::kable(., row.names = F, format="html") %>% 
  kableExtra::kable_styling("striped", full_width = T)
```


### Load RNA-seq data: GSE48731.MF2
```{r message=FALSE}
 # list.files("../source_data/rna_seq/GSE148731")
# rna.seq.data.files <- list()
rna.seq.data.files$GSE148731.MF2 <- "../source_data/rna_seq/GSE148731/GSE148731_toptable_MF2_01.Rmd.2020_07_09.txt"
stopifnot(file.exists(rna.seq.data.files$GSE148731.MF2))
 # rna.seq.data.list <- list()
rna.seq.data.list$GSE148731.MF2 <- read.table(file = rna.seq.data.files$GSE148731.MF2, header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>%
   select(c("Gene.symbol", "logFC")) %>% 
  # mutate(Gene.symbol = toupper(Gene.symbol)) %>% 
  # summarise(logFC = median(logFC), .groups="drop") %>% 
   as.data.frame(., stringsAsFactors=FALSE)
head(rna.seq.data.list$GSE148731.MF2)
```


```{r}
# WITH AGGREGATING:
dim(rna.seq.data.list$GSE148731.MF2) #  17085     2
sapply(rna.seq.data.list$GSE148731.MF2, class) # "character"   "numeric"
range(rna.seq.data.list$GSE148731.MF2$logFC) # -7.524136  9.795100
sum(rna.seq.data.list$GSE148731.MF2$logFC == 0) #
sum(duplicated(rna.seq.data.list$GSE148731.MF2$Gene.symbol)) # 0

```

```{r}
head(rna.seq.data.list$GSE148731.MF2) %>% 
  knitr::kable(., row.names = F, format="html") %>% 
  kableExtra::kable_styling("striped", full_width = T)
```

## Combine all data frames  
```{r results="hide"}
length(rna.seq.data.list) # 4
length(geo.fold.change.list) # 8
all.data.list <- c(geo.fold.change.list, rna.seq.data.list)
stopifnot(all(sapply(all.data.list, class) == "data.frame"))
```

```{r results="hide"}
names(all.data.list)
lapply(all.data.list, names)
length(all.data.list) # 8
```


### Merge data frames  
```{r}
merged.df <- all.data.list %>% purrr::reduce(dplyr::full_join, by="Gene.symbol") %>% 
  tibble::column_to_rownames("Gene.symbol") %>% 
  set_colnames(names(all.data.list))

# dim(merged.df) # 30687    12
head(merged.df) %>% 
  knitr::kable(., row.names = T, format="html") %>% 
  kableExtra::kable_styling("striped", full_width = T)
```

Summary table:  
```{r}
df.summary <- data.frame(
  Column = names(merged.df),
  Min = round(sapply(merged.df, min, na.rm=T), 2),
  Max = round(sapply(merged.df, max, na.rm=T), 2),
  Count.zero = sapply(merged.df, function(x) {sum(x==0, na.rm=T)}),
  Count.NA = comma(sapply(merged.df, function(x) {sum(is.na(x))}))
) %>% 
  mutate(Count.zero = comma(as.integer(Count.zero), accuracy=1))
knitr::kable(df.summary, row.names = F, format="html", align=rep("l", 5)) %>% 
  kableExtra::kable_styling("striped", full_width = T)
```


```{r results="asis"}
cat("Number of rows in merged data frame:<b>", comma(nrow(merged.df)), "</b><br>\n")
cat("Number of gene symbols in merged data frame:<b>", comma(length(unique(row.names(merged.df)))), "</b><br>\n")
```

```{r results="asis"}
cat("Number of complete data rows: <b>", comma(sum(complete.cases(merged.df))), "</b> (",
    percent(sum(complete.cases(merged.df))/length(unique(row.names(merged.df)))),
    ")<br>\n", sep="")
```

## Remove all rows with any missing data  
```{r results="hide"}
max.na <- 0
merged.df.filt.na <- merged.df[complete.cases(merged.df),]
dim(merged.df.filt.na)   # 7,729   12
dim(merged.df)    #  30,687     12
```

```{r results="asis"}
cat("Number of genes remaining in filtered data: <b>",
    comma(nrow(merged.df.filt.na)),
    "</b> (", percent(nrow(merged.df.filt.na) / nrow(merged.df)), ")", sep="")
```

### Summary table after filtering NAs:  
```{r}
df.summary.2 <- data.frame(
  Column = names(merged.df.filt.na),
  Dataset=names(all.data.list),
  Min = round(sapply(merged.df.filt.na, min, na.rm=T), 2),
  Max = round(sapply(merged.df.filt.na, max, na.rm=T), 2),
  Count.zero = sapply(merged.df.filt.na, function(x) {sum(x==0, na.rm=T)}),
  Count.NA = comma(sapply(merged.df.filt.na, function(x) {sum(is.na(x))}))
) %>% 
  mutate(Count.zero = comma(as.integer(Count.zero), accuracy=1))
knitr::kable(df.summary.2, row.names = F, format="html", align=rep("l", 5)) %>% 
  kableExtra::kable_styling("striped", full_width = T)
rm(df.summary.2)
```



## Calculate ranks (1)  

* Convert logFC to ranks  
* ties are averaged  

### Ranks: downregulated  

```{r}
ranks.down.mx <- merged.df.filt.na %>% 
  tibble::rownames_to_column("Gene") %>% 
  dplyr::mutate(across(-Gene, rank, ties.method="average", na.last="keep")) %>%
  dplyr::rowwise() %>% 
  dplyr::mutate(ranksum = sum(c_across(-Gene))) %>% 
  dplyr::arrange(ranksum) %>% 
  dplyr::select(-ranksum) %>% 
  dplyr::ungroup() %>% 
  tibble::column_to_rownames("Gene") %>% 
  as.matrix
stopifnot(all(complete.cases(ranks.down.mx)))
head(ranks.down.mx)
```

```{r results="hide"}
apply(ranks.down.mx, 2, max)
```

### Ranks: upregulated  
```{r}
ranks.up.mx <- as.matrix(merged.df.filt.na) %>% 
  magrittr::multiply_by(-1) %>% 
  as.data.frame(., stringsAsFactors=F) %>%
  tibble::rownames_to_column("Gene") %>% 
  dplyr::mutate(across(-Gene, rank, ties.method="average", na.last="keep")) %>%
  dplyr::rowwise() %>% 
  dplyr::mutate(ranksum = sum(c_across(-Gene))) %>% 
  dplyr::arrange(ranksum) %>% 
  dplyr::select(-ranksum) %>% 
  dplyr::ungroup() %>% 
  tibble::column_to_rownames("Gene") %>% 
  as.matrix
stopifnot(all(complete.cases(ranks.up.mx)))
head(ranks.up.mx)
```

## Plot multidimensional scaling analysis 
(based on Figure S1, Doublet et al., 2017)  

```{r}
dist.mx <- dist(t(ranks.down.mx), method = "manhattan")

# dim(as.matrix(dist.mx)) # 8 8
mds <- cmdscale(dist.mx) %>% 
  as.data.frame(.) %>% 
  set_colnames(c("Dim1", "Dim2")) %>%
  dplyr::mutate(Dataset = names(all.data.list)) %>% 
  dplyr::mutate(Technology = ifelse(Dataset %in% names(rna.seq.data.list), "RNA-seq", "Microarray")) %>%
  dplyr::mutate(Cell.type = ifelse(grepl("_DC", Dataset), "Dendritic cells", "Macrophages")) %>% 
  dplyr::mutate(Cell.type = ifelse(grepl("GSE108363", Dataset), "Whole blood", Cell.type)) %>% 
  dplyr::mutate(Dataset = stringr::str_remove(Dataset, "_.*$")) %>% 
  dplyr::mutate(Dataset.Num = seq_len(nrow(.))) %>% 
  dplyr::select(Dataset.Num, everything())

mds %>% 
  dplyr::mutate(Dim1 = comma(Dim1)) %>% 
  dplyr::mutate(Dim2 = comma(Dim2)) %>% 
  knitr::kable(., row.names = F, format="html") %>% 
  kableExtra::kable_styling("striped")
rm(dist.mx)
```

```{r}
set.seed(2)
font.size <- 10
gg.mds <- ggplot(mds, aes(x=Dim1, y=Dim2, colour=Cell.type, shape=Technology, label=Dataset.Num)) +
  geom_point(size=5) +
  # geom_text() +
  geom_text_repel(colour="black", fontface="bold", box.padding=0.5, min.segment.length=10) +
  xlab("Dimension 1") + ylab("Dimension 2") + 
  geom_vline(xintercept = 0, colour="darkgrey") +
  geom_hline(yintercept = 0, colour="darkgrey") +
  theme(axis.text=element_text(size=font.size, color="black")) +
  theme(axis.title=element_text(size=font.size, color="black", face="bold")) +
  theme(legend.text = element_text(size=font.size, face="bold")) +
  theme(legend.title = element_text(size=font.size, face="bold")) + 
  scale_shape_manual(values=c(18, 19))
gg.mds
```

### Save plot as pdf  
```{r}
out.file <- paste("data_out", stringr::str_replace(this.script, "\\.[Rr][Mm][Dd]$", ".pdf"), sep="/")
ggsave(gg.mds, filename = out.file, height = 10, width = 17, units = "cm", useDingbats=)
out.file
```



## Session info  
<button class="button" onclick="myFunction('DIV_5')">Show/hide session info</button>
<div id="DIV_5" class="div_default_hide">

```{r print_date_and_time}
Sys.time()
```

```{r print_session_info}
# sessionInfo()
devtools::session_info()
```
</div>
  
<script>
  var divsToHide = document.getElementsByClassName("div_default_hide");
for(var i = 0; i < divsToHide.length; i++)
{
  divsToHide[i].style.display = 'none';
}
</script>
